### 2024-03-11 paper.tex
Update on Overleaf.

\documentclass[sigconf]{acmart}

%% Rights management information.
\setcopyright{acmlicensed}
\copyrightyear{2024}
\acmYear{2024}
\acmDOI{XXXXXXX.XXXXXXX}
\acmSubmissionID{123-A56-BU3}

\begin{document}
\title{Memory Safety: A Systematic Literature Review}

\author{Shane Panter}
    \affiliation{%
    \institution{Boise State University}
    \city{Boise}
    \state{Idaho}
    \country{USA}}
    \email{shanepanter@boisestate.edu}

\author{Nasir U. Eisty}
    \affiliation{%
    \institution{Boise State University}
    \city{Boise}
    \state{Idaho}
    \country{USA}}
    \email{nasireisty@boisestate.edu}

\renewcommand{\shortauthors}{Panter et al.}

\begin{abstract}
    \textit{Context:}
    TODO:
    \textit{Objective:}
    TODO:
    \textit{Method:}
    TODO:
    \textit{Results:}
    TODO:
    \textit{Conclusions:}
\end{abstract}

%%
%% The code below is generated by the tool at http://dl.acm.org/ccs.cfm.
%% Please copy and paste the code instead of the example below.
%%
\begin{CCSXML}
<ccs2012>
 <concept>
  <concept_id>00000000.0000000.0000000</concept_id>
  <concept_desc>Do Not Use This Code, Generate the Correct Terms for Your Paper</concept_desc>
  <concept_significance>500</concept_significance>
 </concept>
 <concept>
  <concept_id>00000000.00000000.00000000</concept_id>
  <concept_desc>Do Not Use This Code, Generate the Correct Terms for Your Paper</concept_desc>
  <concept_significance>300</concept_significance>
 </concept>
 <concept>
  <concept_id>00000000.00000000.00000000</concept_id>
  <concept_desc>Do Not Use This Code, Generate the Correct Terms for Your Paper</concept_desc>
  <concept_significance>100</concept_significance>
 </concept>
 <concept>
  <concept_id>00000000.00000000.00000000</concept_id>
  <concept_desc>Do Not Use This Code, Generate the Correct Terms for Your Paper</concept_desc>
  <concept_significance>100</concept_significance>
 </concept>
</ccs2012>
\end{CCSXML}

\ccsdesc[500]{Do Not Use This Code~Generate the Correct Terms for Your Paper}
\ccsdesc[300]{Do Not Use This Code~Generate the Correct Terms for Your Paper}
\ccsdesc{Do Not Use This Code~Generate the Correct Terms for Your Paper}
\ccsdesc[100]{Do Not Use This Code~Generate the Correct Terms for Your Paper}

%%
%% Keywords. The author(s) should pick words that accurately describe
%% the work being presented. Separate the keywords with commas.
\keywords{Memory safety}

\received{20 February 2007}
\received[revised]{12 March 2009}
\received[accepted]{5 June 2009}

\maketitle
\section{Introduction}
TODO~\cite{Stuart2007-cc}

\section{Research Methodology}

\subsection{Planning}

\subsubsection{Research Questions}

\begin{itemize}
    \item RQ1: 
    \item RQ2:
    \item RQ3:
\end{itemize}

\subsubsection{Search Strategy:}

\subsubsection{Search criteria:}

\subsubsection{Inclusion and exclusion criteria:}

\subsection{Conducting the review}

\subsubsection{Study search and selection:}

\subsubsection{Data extraction:}

\subsubsection{Data synthesis:}

\section{Results}

\subsection{RQ1}
\subsection{RQ2}
\subsection{RQ3}

\section{Conclusion}

\section{Threats to Validity}

\bibliographystyle{ACM-Reference-Format}
\bibliography{paperpile.bib}

\end{document}

### 2024-03-13 paper.tex
Added makefile and .gitignore
Starting on introduction

@@
 \maketitle
+
 \section{Introduction}
-TODO~\cite{Stuart2007-cc}
+
+The 1995 movie Hackers~\cite{Wikipedia_contributors2024-zr} prescient predictions regarding the ease
+of breaking into computing systems in cyberspace have came to fruition. The White House Office of
+the National Cyber Director (ONCD) released a report calling for the technical community to
+proactively reduce the attach surface in cyberspace with a two pronged approach. First we need to
+address the root cause of many of the most heinous cyber attacks, memory unsafe programming
+languages~\cite{United_States_Gov2024-pp}. Second, we need to establish better cybersecurity quality
+metrics so we can have a better understanding of the cyber security landscape.
+
+In the ever-evolving landscape of software development, the reliability and security of computer
+systems stand as paramount concerns for all parties involved. Modern software is constructed by
+building ever more complex abstractions. Each abstraction forms a layer that the next one it built
+upon. Thus, if we aim to have a secure system we must peel back all the layers and tackle one of the
+fundamental abstractions in computer science the programming language. Programming languages that
+provided and enforce memory safety eliminate whole classes of bugs such as buffer overflows,
+dangling pointers, and memory leaks which have been implicated in a myriad of security
+vulnerabilities and system crashes.
+
+This Systematic Literature Review aims to provide a comprehensive and up-to-date analysis of the
+existing body of research surrounding memory safety. By systematically reviewing and synthesizing
+the wealth of knowledge accumulated in this field, we seek to uncover the current state of
+understanding, identify emerging trends, and highlight gaps in our knowledge that warrant further
+investigation.
 
 \section{Research Methodology}
+
+For our research methodolgy we ~\cite{Stuart2007-cc}
 
@@
 \section{Threats to Validity}
+\balance
 

### 2024-03-13 paper.tex
Working on defining research questions

@@
 \documentclass[sigconf]{acmart}
+\usepackage{listings, listings-rust}
+\usepackage{xcolor}
+
+\NewDocumentCommand{\codeword}{v}{%
+\texttt{\textcolor{blue}{#1}}
+}
+
+\lstset{language=Rust, keywordstyle={\bfseries \color{blue}}}
 
@@
 the National Cyber Director (ONCD) released a report calling for the technical community to
-proactively reduce the attach surface in cyberspace with a two pronged approach. First we need to
+proactively reduce the attack surface in cyberspace with a two pronged approach. First we need to
 address the root cause of many of the most heinous cyber attacks, memory unsafe programming
@@
 
-This Systematic Literature Review aims to provide a comprehensive and up-to-date analysis of the
-existing body of research surrounding memory safety. By systematically reviewing and synthesizing
-the wealth of knowledge accumulated in this field, we seek to uncover the current state of
-understanding, identify emerging trends, and highlight gaps in our knowledge that warrant further
+This Systematic Literature Review (SLR) aims to provide a comprehensive and up-to-date analysis of
+the existing body of research surrounding memory safety. By systematically reviewing and
+synthesizing the wealth of knowledge accumulated in this field, we seek to uncover the current state
+of understanding, identify emerging trends, and highlight gaps in our knowledge that warrant further
 investigation.
@@
 
-For our research methodolgy we ~\cite{Stuart2007-cc}
+For our research methodolgy we followed the advice of Kitchenham and Charters~\cite{Stuart2007-cc}
+and divided our review into the 3 discrete phases, planning the review, conducting the review, and
+reporting the review results.
 
 \subsection{Planning}
+
+Before we start researching we must first confirm the need for a SLR. The recent report released by
+ONCD~\cite{United_States_Gov2024-pp} has conveniently done this job for us by compiling a report
+detailing the need for research in the domain of memory safety. Thus, we must first define the most
+important aspect of the planning phase, the research questions.
 
@@
 \begin{itemize}
-    \item RQ1:
-    \item RQ2:
-    \item RQ3:
+    \item \textbf{RQ1:} How can \lstinline{unsafe} Rust derived from C be converted to safe Rust?
+    \item \textbf{RQ2:} How can we build Rust bindings into core abstractions?
+    \item \textbf{RQ3:} How can we better find vulnerabilities in \lstinline{unsafe} blocks
 \end{itemize}
@@
 
+In this section we address the proposed research questions.
+
 \subsection{RQ1}

### 2024-03-14 paper.tex
First draft of RQ

@@
 ONCD~\cite{United_States_Gov2024-pp} has conveniently done this job for us by compiling a report
-detailing the need for research in the domain of memory safety. Thus, we must first define the most
-important aspect of the planning phase, the research questions.
+detailing the need for research in the domain of memory safety. While the ONCD report detailed a two
+pronged approach for this SLR we will be focusing on a memory safe programming language,
+specifically Rust.
 
 \subsubsection{Research Questions}
+We have defined the following research questions to drive this study.
 

### 2024-04-01 paper.tex
Change paperpile bib to export based on tag not folder
Working on SLR protocal section

@@
 \usepackage{xcolor}
+\pagestyle{plain} %to show page numbers
 
@@
 In the ever-evolving landscape of software development, the reliability and security of computer
-systems stand as paramount concerns for all parties involved. Modern software is constructed by
-building ever more complex abstractions. Each abstraction forms a layer that the next one it built
-upon. Thus, if we aim to have a secure system we must peel back all the layers and tackle one of the
-fundamental abstractions in computer science the programming language. Programming languages that
-provided and enforce memory safety eliminate whole classes of bugs such as buffer overflows,
-dangling pointers, and memory leaks which have been implicated in a myriad of security
-vulnerabilities and system crashes.
+systems stands as a paramount concern for all parties involved. Modern software is constructed by
+building ever more complex abstractions, one on top of the other. Thus, if we aim to have a secure
+system we must start to peel back all the layers and tackle one of the fundamental abstractions in
+computer science, the programming language. Programming languages that provided and enforce memory
+safety eliminate whole classes of bugs such as buffer overflows, dangling pointers, and memory leaks
+which have been implicated in a myriad of security vulnerabilities and system crashes.
 
-This Systematic Literature Review (SLR) aims to provide a comprehensive and up-to-date analysis of
-the existing body of research surrounding memory safety. By systematically reviewing and
-synthesizing the wealth of knowledge accumulated in this field, we seek to uncover the current state
-of understanding, identify emerging trends, and highlight gaps in our knowledge that warrant further
-investigation.
+The migration of legacy C codebases to memory-safe programming languages like Rust is of paramount
+importance in modern software development, aiming to mitigate common security vulnerabilities and
+memory-related errors. This paper presents a systematic literature review (SLR) focusing on
+strategies and methodologies for converting C code to Rust while minimizing the utilization of
+unsafe blocks. Our study aims to provide a comprehensive overview of existing research, identify
+gaps, and suggest future directions in this domain. Through a rigorous search process, we
+synthesized relevant studies and extracted key findings to offer insights into effective approaches
+for ensuring memory safety during code migration.
 
@@
 pronged approach for this SLR we will be focusing on a memory safe programming language,
-specifically Rust.
+specifically Rust. While there are many memory safe programming languages available for software
+developers to use Rust is the only language that has support in kernel
+space~\cite{The_kernel_development_community_undated-iw} thus is a primary candidate to replace the
+aging C programming language.
 
@@
 \begin{itemize}
-    \item \textbf{RQ1:} How can \lstinline{unsafe} Rust derived from C be converted to safe Rust?
-    \item \textbf{RQ2:} How can we build Rust bindings into core abstractions?
-    \item \textbf{RQ3:} How can we better find vulnerabilities in \lstinline{unsafe} blocks
+
+    \item \textbf{RQ1:} What are the existing methodologies and strategies for converting C code to
+      Rust?
+    \item \textbf{RQ2:} How do these methodologies address memory safety concerns and minimize the
+      use of \lstinline{unsafe} blocks?
+    \item \textbf{RQ3:} What are the challenges and limitations encountered in the conversion process?
 \end{itemize}
@@
 
+We employed a robust multi-step search strategy across three digital databases, ACM Digital Library,
+IEEE Xplore, and Google Scholar in order to find all the current research that has been done. These
+databases cover a wide range of current software engineering and computer science literature. We
+leveraged the advanced search features of both the ACM Digital Library and IEEE Xplore to search the
+titles and abstracts for keywords using boolean search operators. We then searched Google Scholar
+using the same keywords and operators sorted by relevance and took the top 250 results.
+
 \subsubsection{Search criteria:}
 
+The search criteria for each database is shown in table~\ref{tab:keywords}.  First, we searched
+titles and abstracts in the ACM Digital Library and IEEE Xplore with the keywords ``C rust convert
+unsafe'' from May 2015 to the present~\footnote{This was the release date of rust 1.0}. Then we
+searched on google Google Scholar using the same keywords and dates in order to to cast as wide a
+net as possible.
+
+\begin{table}
+\begin{tabular}{|| p{3cm}| p{5cm} ||}
+ \hline
+ Database & Query \\
+ \hline\hline
+ ACM Digital Library & Title:(C conver* unsafe rust) AND Abstract:(C conver* unsafe rust)  \\
+ IEEE Xplore & TBD \\
+ Google Scholar & TBD \\
+ \hline
+\end{tabular}
+\caption{Search Queries used for each Database.}
+\label{tab:keywords}
+\end{table}
+
+Search keywords: "C to Rust conversion", "memory-safe Rust", "legacy code migration", "unsafe blocks
+mitigation", "safe programming practices".
+
 \subsubsection{Inclusion and exclusion criteria:}
+
+Inclusion criteria: Papers published between 2012 and 2024, written in English, focusing on
+converting C code to Rust with a focus on memory safety.
+
+Exclusion criteria: Duplicate studies, non-peer-reviewed articles, unrelated topics.
+
+Manual search: Relevant conference proceedings, journals, and citations of selected papers.
 
@@
 
+Initial screening based on titles and abstracts.  Full-text assessment for eligibility based on
+inclusion and exclusion criteria.  Selection of final studies for data extraction and analysis.
+
 \subsubsection{Data extraction:}
+
+Extraction of relevant data including methodologies, tools, challenges, and outcomes.
 
@@
 
+Synthesis of key findings and categorization based on common themes.
+Identification of trends, gaps, and areas for further investigation.
+
 \section{Results}
 
-In this section we address the proposed research questions.
+Presentation of synthesized findings on existing methodologies and strategies.  Discussion on the
+effectiveness of approaches in ensuring memory safety and minimizing unsafe blocks.  Analysis of
+challenges and limitations encountered in the conversion process.  Comparison of different tools and
+techniques used in C to Rust migration.
+
+Implications of the findings for software developers, researchers, and practitioners.
+Recommendations for improving existing methodologies and addressing identified gaps.
+Suggestions for future research directions in the field of C to Rust conversion.
 
@@
 
+Summary of key findings and insights obtained from the systematic literature review.
+Concluding remarks on the significance of memory-safe programming languages in mitigating security vulnerabilities.
+Closing thoughts on the importance of minimizing unsafe blocks during code migration.
+
 \section{Threats to Validity}
@@
 \bibliographystyle{ACM-Reference-Format}
-\bibliography{paperpile.bib}
+\bibliography{slr-paper-memory-safety-rust.bib}
 

### 2024-04-10 paper.tex
Working on slr paper

@@
 \usepackage{xcolor}
-\pagestyle{plain} %to show page numbers
+\usepackage[raggedrightboxes]{ragged2e}
 
@@
 \begin{document}
-\title{Memory Safety: A Systematic Literature Review}
+\title{Rust for Operating System Development: A Systematic Literature Review}
 
-\author{Shane Panter}
-    \affiliation{%
-    \institution{Boise State University}
-    \city{Boise}
-    \state{Idaho}
-    \country{USA}}
-    \email{shanepanter@boisestate.edu}
+%% \author{Shane Panter}
+%%     \affiliation{%
+%%     \institution{Boise State University}
+%%     \city{Boise}
+%%     \state{Idaho}
+%%     \country{USA}}
+%%     \email{shanepanter@boisestate.edu}
 
-\author{Nasir U. Eisty}
-    \affiliation{%
-    \institution{Boise State University}
-    \city{Boise}
-    \state{Idaho}
-    \country{USA}}
-    \email{nasireisty@boisestate.edu}
+%% \author{Nasir U. Eisty}
+%%     \affiliation{%
+%%     \institution{Boise State University}
+%%     \city{Boise}
+%%     \state{Idaho}
+%%     \country{USA}}
+%%     \email{nasireisty@boisestate.edu}
 
@@
 memory-related errors. This paper presents a systematic literature review (SLR) focusing on
-strategies and methodologies for converting C code to Rust while minimizing the utilization of
-unsafe blocks. Our study aims to provide a comprehensive overview of existing research, identify
-gaps, and suggest future directions in this domain. Through a rigorous search process, we
-synthesized relevant studies and extracted key findings to offer insights into effective approaches
-for ensuring memory safety during code migration.
+strategies and methodologies for integrating Rust into one of the most fundamental C codebases in
+the world, the Linux Kernel.  Our study aims to provide a comprehensive overview of existing
+research, identify gaps, and suggest future directions in this domain. Through a rigorous search
+process, we synthesized relevant studies and extracted key findings to offer insights into effective
+approaches for ensuring memory safety during code migration.
 
@@
 \subsubsection{Research Questions}
+\label{sec:researchQuestions}
+
 We have defined the following research questions to drive this study.
@@
 \begin{itemize}
+    \item \textbf{RQ1:} What are the existing approaches, methodologies, and best practices for
+      implementing operating system kernels in Rust?
+    \item \textbf{RQ2:} What are the performance implications of using Rust for operating system
+      kernel development in terms of throughput, latency, and resource utilization?
+    \item \textbf{RQ3:} What are the major challenges and limitations encountered when developing
+      operating system kernels in Rust, particularly in comparison to other languages?
+\end{itemize}
 
-    \item \textbf{RQ1:} What are the existing methodologies and strategies for converting C code to
-      Rust?
-    \item \textbf{RQ2:} How do these methodologies address memory safety concerns and minimize the
-      use of \lstinline{unsafe} blocks?
-    \item \textbf{RQ3:} What are the challenges and limitations encountered in the conversion process?
-\end{itemize}
+%% Other RQ's
+%% \item What are the key motivations and advantages for developing operating system kernels in Rust
+%% compared to traditional languages like C or C++?
+%% \item What are the fundamental design principles and features of Rust that make it suitable for
+%% building operating system kernels?
+%% \item What are the existing approaches, methodologies, and best practices for implementing operating
+%% system kernels in Rust?
+%% \item What are the major challenges and limitations encountered when developing operating system
+%% kernels in Rust, particularly in comparison to other languages?
+%% \item How does the safety, security, and reliability of operating system kernels developed in Rust
+%% compare to those developed using other programming languages?
+%% \item What are the performance implications of using Rust for operating system kernel development in
+%% terms of throughput, latency, and resource utilization?
+%% \item What is the current state of adoption and usage of Rust in real-world operating system kernel
+%% projects, and what are the lessons learned from these experiences?
+%% \item How does the Rust ecosystem support operating system kernel development, including libraries,
+%% tools, and community resources?
+%% \item What are the potential future directions and research opportunities for improving the development
+%% of operating system kernels in Rust?
+%% \item What are the implications of using Rust for operating system kernel development in terms of
+%% maintenance, extensibility, and compatibility with existing hardware and software architectures?
 
@@
 leveraged the advanced search features of both the ACM Digital Library and IEEE Xplore to search the
-titles and abstracts for keywords using boolean search operators. We then searched Google Scholar
-using the same keywords and operators sorted by relevance and took the top 250 results.
+titles and abstracts for keywords using boolean search operators. The ACM Digital Library and IEEE
+Xplore formed the core of our keyword search and we used Google Scholar for backward and forward
+snowballing.
 
@@
 
-The search criteria for each database is shown in table~\ref{tab:keywords}.  First, we searched
-titles and abstracts in the ACM Digital Library and IEEE Xplore with the keywords ``C rust convert
-unsafe'' from May 2015 to the present~\footnote{This was the release date of rust 1.0}. Then we
-searched on google Google Scholar using the same keywords and dates in order to to cast as wide a
-net as possible.
+We first searched abstracts in the ACM Digital Library and metadata in IEEE Xplore with the keywords
+listed in table~\ref{tab:keywords} between the years 2019~\cite{Gaynor2019-dk} and 2024. We then
+used Google Scholar for forward and backward snowballing.
 
 \begin{table}
-\begin{tabular}{|| p{3cm}| p{5cm} ||}
+\begin{tabular}{|| p{2cm}| p{4cm} | p{1cm} ||}
  \hline
- Database & Query \\
+ Database & Query & Results \\
  \hline\hline
- ACM Digital Library & Title:(C conver* unsafe rust) AND Abstract:(C conver* unsafe rust)  \\
- IEEE Xplore & TBD \\
- Google Scholar & TBD \\
+ ACM  & Abstract:("operating system" kernel linux) AND Abstract:(rust) & 22 \\
+ IEEE & ("All Metadata":"operating system" AND "All Metadata":rust) OR ("All Metadata":kernel AND
+ "All Metadata":rust) OR ("All Metadata":linux AND "All Metadata":rust) & 37 \\
  \hline
@@
 
-Search keywords: "C to Rust conversion", "memory-safe Rust", "legacy code migration", "unsafe blocks
-mitigation", "safe programming practices".
-
 \subsubsection{Inclusion and exclusion criteria:}
 
-Inclusion criteria: Papers published between 2012 and 2024, written in English, focusing on
-converting C code to Rust with a focus on memory safety.
-
-Exclusion criteria: Duplicate studies, non-peer-reviewed articles, unrelated topics.
-
-Manual search: Relevant conference proceedings, journals, and citations of selected papers.
+Following the guidelines outlined by Kitchenham and Charters~\cite{Stuart2007-cc} we set the
+inclusion and exclusion criteria based on our research questions outlined in
+section~\ref{sec:researchQuestions}. We only consider papers that are written in English, published
+in conferences or journals, and within the time frame of 2019 and 2024. The published papers should
+describe using the Rust programming language for either developing a new kernel, extending an
+existing kernel, or authoring drivers. We will include any type of kernel architecture including
+monolithic, micro, or uni as long as the paper is using the Rust programming language.
 

### 2024-04-17 paper.tex
Updated RQ

@@
 \begin{itemize}
-    \item \textbf{RQ1:} What are the existing approaches, methodologies, and best practices for
-      implementing operating system kernels in Rust?
+    \item \textbf{RQ1:} What are the existing approaches and methodologies for implementing
+      operating system kernels in Rust?
     \item \textbf{RQ2:} What are the performance implications of using Rust for operating system
@@
       operating system kernels in Rust, particularly in comparison to other languages?
+    \item \textbf{RQ4:} What are the best practices for implementing operating system kernels in Rust?
 \end{itemize}

### 2024-04-22 paper.tex
Updated search strings for Scopus database

@@
 system we must start to peel back all the layers and tackle one of the fundamental abstractions in
-computer science, the programming language. Programming languages that provided and enforce memory
+computer science, the programming language. Programming languages that provide and enforce memory
 safety eliminate whole classes of bugs such as buffer overflows, dangling pointers, and memory leaks
@@
 detailing the need for research in the domain of memory safety. While the ONCD report detailed a two
-pronged approach for this SLR we will be focusing on a memory safe programming language,
+pronged approach, for this SLR we will be focusing on a memory safe programming language,
 specifically Rust. While there are many memory safe programming languages available for software
@@
 We employed a robust multi-step search strategy across three digital databases, ACM Digital Library,
-IEEE Xplore, and Google Scholar in order to find all the current research that has been done. These
+IEEE Xplore, and Scopus in order to find all the current research that has been done. These
 databases cover a wide range of current software engineering and computer science literature. We
-leveraged the advanced search features of both the ACM Digital Library and IEEE Xplore to search the
-titles and abstracts for keywords using boolean search operators. The ACM Digital Library and IEEE
-Xplore formed the core of our keyword search and we used Google Scholar for backward and forward
-snowballing.
+leveraged the advanced search features of all three databases to search the titles and abstracts for
+keywords using boolean search operators.
 
@@
 
-We first searched abstracts in the ACM Digital Library and metadata in IEEE Xplore with the keywords
-listed in table~\ref{tab:keywords} between the years 2019~\cite{Gaynor2019-dk} and 2024. We then
-used Google Scholar for forward and backward snowballing.
+We search all three databases with the keywords listed in table~\ref{tab:keywords} between January
+1, 2019~\cite{Gaynor2019-dk} and April 1, 2024. While both ACM and IEEE had mutually exclusive
+results Scopus had 44 duplicates that needed to be removed. We then used Google Scholar for forward
+and backward snowballing.
 
@@
  \hline\hline
- ACM  & Abstract:("operating system" kernel linux) AND Abstract:(rust) & 22 \\
+ ACM  & (Abstract:("operating system" kernel linux) AND Abstract:(rust) )OR (Title:("operating
+ system" kernel linux) AND Title:(rust)) & 25 \\
  IEEE & ("All Metadata":"operating system" AND "All Metadata":rust) OR ("All Metadata":kernel AND
- "All Metadata":rust) OR ("All Metadata":linux AND "All Metadata":rust) & 37 \\
+ "All Metadata":rust) OR ("All Metadata":linux AND "All Metadata":rust) & 39 \\
+ Scopus & TITLE-ABS-KEY ( ( "operating system" OR kernel OR linux ) AND rust ) AND PUBYEAR > 2018
+ AND PUBYEAR < 2025 AND ( LIMIT-TO ( SUBJAREA , "COMP" ) ) AND ( LIMIT-TO ( LANGUAGE , "English" ) )
+ & 87 \\
  \hline

### 2024-05-03 paper.tex
Update on Overleaf.
Added cttc for paper review

@@
 \begin{document}
-\title{Rust for Operating System Development: A Systematic Literature Review}
+\title{Kernel Development Using Rust: A Systematic Literature Review}
+%\title{}
 
@@
 existing kernel, or authoring drivers. We will include any type of kernel architecture including
-monolithic, micro, or uni as long as the paper is using the Rust programming language.
+monolithic kernel, microkernel, or unikernel in both the embedded and non-embedded space as long as
+the paper is using the Rust programming language in some way.
 
@@
 
-\subsection{RQ1}
-\subsection{RQ2}
-\subsection{RQ3}
+\begin{table}
+  \begin{tabular}{||c|c||}
+    \hline
+    System Type & Papers\\
+    \hline\hline
+    Embedded & ~\cite{Ayers2022-vk} \\
+    Non-Embedded & 0 \\
+    Unikernel & 0 \\
+    Kernel & 0 \\
+    \hline
+  \end{tabular}
+\end{table}
+
+This is a list of operating systems.
+\begin{itemize}
+\item \href{https://docs.kernel.org/rust/}{Linux kernel v6.1+}
+\item \href{https://github.com/tock/tock}{Tock}
+\item \href{https://hubris.oxide.computer/}{Hubris OS}
+\item Theseus
+\end{itemize}
 

### 2024-05-03 paper.tex
Working on reading all papers

@@
  \hline\hline
- ACM  & (Abstract:("operating system" kernel linux) AND Abstract:(rust) )OR (Title:("operating
- system" kernel linux) AND Title:(rust)) & 25 \\
+ ACM  & (Abstract:("operating system" kernel linux OS) AND Abstract:(rust) )OR (Title:("operating
+ system" OS kernel linux) AND Title:(rust))  & 30 \\
  IEEE & ("All Metadata":"operating system" AND "All Metadata":rust) OR ("All Metadata":kernel AND

### 2024-05-17 paper.tex
Working on reading through papers
Working on Paper
Adding in all cttc summaries thus far.

@@
 \usepackage[raggedrightboxes]{ragged2e}
+\pagestyle{plain}
 
@@
 memory-related errors. This paper presents a systematic literature review (SLR) focusing on
-strategies and methodologies for integrating Rust into one of the most fundamental C codebases in
-the world, the Linux Kernel.  Our study aims to provide a comprehensive overview of existing
-research, identify gaps, and suggest future directions in this domain. Through a rigorous search
-process, we synthesized relevant studies and extracted key findings to offer insights into effective
-approaches for ensuring memory safety during code migration.
+strategies and methodologies for integrating Rust into one of the most fundamental areas that are
+typically dominated by unsafe languages the operating system kernel. Our study aims to provide a
+comprehensive overview of existing research, identify gaps, and suggest future directions in this
+domain. Through a rigorous search process, we synthesized relevant studies and extracted key
+findings to offer insights into effective approaches for ensuring memory safety when working closley
+with hardware.
 
@@
 and divided our review into the 3 discrete phases, planning the review, conducting the review, and
-reporting the review results.
+reporting the review results. The following sections detail our review process.
 
@@
 pronged approach, for this SLR we will be focusing on a memory safe programming language,
-specifically Rust. While there are many memory safe programming languages available for software
-developers to use Rust is the only language that has support in kernel
-space~\cite{The_kernel_development_community_undated-iw} thus is a primary candidate to replace the
-aging C programming language.
+specifically Rust. While there are many modern memory safe programming languages available for software
+developers to use Rust is one of the few languages that is feasible to use when developing operating
+system kernels due to is lack of runtime and garbage collector. Even the Linux kernel is in the
+early stages of adding real support for Rust~\cite{The_kernel_development_community_undated-iw}
+thus, we will focus on Rust as a primary candidate to replace the aging C programming language.
 
@@
       kernel development in terms of throughput, latency, and resource utilization?
-    \item \textbf{RQ3:} What are the major challenges and limitations encountered when developing
-      operating system kernels in Rust, particularly in comparison to other languages?
-    \item \textbf{RQ4:} What are the best practices for implementing operating system kernels in Rust?
+    \item \textbf{RQ3:} What are the major challenges, limitations, and lessons learned when
+      developing operating system kernels in Rust, particularly in comparison to other languages?
 \end{itemize}
-
-%% Other RQ's
-%% \item What are the key motivations and advantages for developing operating system kernels in Rust
-%% compared to traditional languages like C or C++?
-%% \item What are the fundamental design principles and features of Rust that make it suitable for
-%% building operating system kernels?
-%% \item What are the existing approaches, methodologies, and best practices for implementing operating
-%% system kernels in Rust?
-%% \item What are the major challenges and limitations encountered when developing operating system
-%% kernels in Rust, particularly in comparison to other languages?
-%% \item How does the safety, security, and reliability of operating system kernels developed in Rust
-%% compare to those developed using other programming languages?
-%% \item What are the performance implications of using Rust for operating system kernel development in
-%% terms of throughput, latency, and resource utilization?
-%% \item What is the current state of adoption and usage of Rust in real-world operating system kernel
-%% projects, and what are the lessons learned from these experiences?
-%% \item How does the Rust ecosystem support operating system kernel development, including libraries,
-%% tools, and community resources?
-%% \item What are the potential future directions and research opportunities for improving the development
-%% of operating system kernels in Rust?
-%% \item What are the implications of using Rust for operating system kernel development in terms of
-%% maintenance, extensibility, and compatibility with existing hardware and software architectures?
 
@@
 We search all three databases with the keywords listed in table~\ref{tab:keywords} between January
-1, 2019~\cite{Gaynor2019-dk} and April 1, 2024. While both ACM and IEEE had mutually exclusive
-results Scopus had 44 duplicates that needed to be removed. We then used Google Scholar for forward
-and backward snowballing.
+1, 2019 and April 1, 2024. While both ACM and IEEE had mutually exclusive results Scopus had 44
+duplicates that needed to be removed. We then used Google Scholar for forward and backward
+snowballing.
 
@@
 monolithic kernel, microkernel, or unikernel in both the embedded and non-embedded space as long as
-the paper is using the Rust programming language in some way.
+the paper is using the Rust programming language in some way. Papers that describe solutions that
+reside 100\% in user space will be excluded.
 
@@
 
-\subsubsection{Study search and selection:}
-
-Initial screening based on titles and abstracts.  Full-text assessment for eligibility based on
-inclusion and exclusion criteria.  Selection of final studies for data extraction and analysis.
-
-\subsubsection{Data extraction:}
-
-Extraction of relevant data including methodologies, tools, challenges, and outcomes.
-
-\subsubsection{Data synthesis:}
-
-Synthesis of key findings and categorization based on common themes.
-Identification of trends, gaps, and areas for further investigation.
+Once we had our initial collection of papers from the databases we merged all duplicate
+papers into one record and then started our first pass which consisted of reading the title and
+abstract and either marking the paper as \textit{include} or \textit{exclude}.
+Once  the initial screening was completed we were left with 32 potential papers that
+we needed an in depth reading.
 
@@
 
-Presentation of synthesized findings on existing methodologies and strategies.  Discussion on the
-effectiveness of approaches in ensuring memory safety and minimizing unsafe blocks.  Analysis of
-challenges and limitations encountered in the conversion process.  Comparison of different tools and
-techniques used in C to Rust migration.
+In this section we present our findings. We address each research question from RQ1 to RQ3.
 
-Implications of the findings for software developers, researchers, and practitioners.
-Recommendations for improving existing methodologies and addressing identified gaps.
-Suggestions for future research directions in the field of C to Rust conversion.
+\subsection{RQ1}
+
+Boos et al.~\cite{Boos2020-zh} is about an experimental operating system that is aimed at addressing
+the state spill problem. State spill happens when a single service in an operating system can harbor
+a state change, induced by interacting with other services, that can eventually cause a system crash
+or system instability. The operating system is built in Rust and leverages the properties of Rust to
+construct a safer system. Theseus operates in a single address space and single privilege level and
+uses properties of the Rust programming language to realize isolation instead of relying on
+hardware. The design of the OS uses a novel cell based structure where ownership of memory and
+resources is enforced by the compiler. In short the Theseus OS is an operating system that was
+designed to match the Rust language instead of the more traditional route of matching the hardware.
+
+Chen X et al.~\cite{Chen2023-wb} primary focused on developing a micro-kernel in Rust that could be
+formally verified by using both the liner type system or rust and a SMT solver. The new micro-kernel
+named Atmosphere follows the tradition of other micro-kernels by pushing most kernel functionally to
+users-space thus limiting the surface area that needs to be proved. The authors were able to get a
+7.5:1 proof-to-code ratio which is higher that other formerly verified microkernels SeL4 and
+CeriKOS, which both have proof-tocode ratio of 19:1 and 20:1 respectively.
+
+Lankes et al.~\cite{Lankes2019-cm} detailed their experiences with developing a Unikernel in
+Rust. Originally written in C and called HermitCore the authors ported the Kernel to the Rust
+programming language in order to take advantage of Rust's memory safe features. The new Unikernel
+consists of only 3.27\% unsafe Rust with the rest of the code base consisting of safe Rust
+
+\subsection{RQ2}
+
+Culic et al.~\cite{Culic2022-bk} looked at latency issues in the Tock OS. Tock is a new operating
+system that is designed to run on embedded systems but does not provide Real time capabilities. The
+authors attempted to add real time capabilities by integrating eBPF into the Tock kernel to improve
+the interrupt handlers response time. The Authors found that early work (still in the prototype
+stage) lowers the response times of the system and makes interrupt response times 3x.
+
+Gonzalez et al.~\cite{Gonzalez2023-ek} explored using the Rust for Linux Project to implement a
+native UDP driver in Rust in order to explore the performance the Rust programming language. The
+authors were able to get a basic driver working with performance only slightly slower than C using
+the Rust for Linux (RFL) project. The RFL project is still to immature to get a full driver up and
+running but is at a stage were we researchers can start experimenting with different
+approaches.
+
+Li et al.~\cite{Li2019-ru} explored the feasibility of using Rust in kernel space. The authors took
+an existing component, the Out of Memory (OOM) and implemented a replacement using the Rust
+programming language. The non-encapsulated interface Rust component which was almost identical to
+the original C component only introduced a 0.7\% overhead. The encapsulated Rust component on the
+other hand added a 3\% performance overhead. Another important aspect of the kernel is size, we must
+be careful to keep the size of the Rust component as close as we can to the C component so we can
+still run on all the same hardware. The authors found that there was only a 0.06\% size increase
+when compared to the original C implementation.
+
+\subsection{RQ3}
+
+Ayers et al.~\cite{Ayers2022-sf} focused on the size of rust binaries in the embedded
+environment. Their work primary focused on the size of binaries produced using the embedded OS
+Tock. The authors identified four major causes of binary growth:
+
+\begin{itemize}
+
+\item Deeply ingrained monomorphization
+\item Suboptimal compiler generated support code
+\item Hidden data structures and data
+\item Fewer compiler optimizations
+\end{itemize}
+
+In addition to the identified causes of binary growth the Ayers et al.~\cite{Ayers2022-sf} have the
+following 5 recommendations when using Rust in a size constrained environment:
+\begin{itemize}
+  \item Minimize Length + Instantiations of Generic Code
+  \item Use Trait Objects Sparingly
+  \item Don't Panic
+  \item Carefully Use Compiler Generated Support Code
+  \item Don't use static mut
+\end{itemize}
+
+Burtsev et al.~\cite{Burtsev2021-mh} explores what is missing in the Rust programming language to
+help solve the isolation problem in an operating system. Currently Rust lacks the ability to express
+isolation in the heap without external support. The RedLeaf experiential operating system relied on
+an complex interface definition language (IDL) to enforce isolated heaps. The paper enumerates
+several properties of the Rust Language that could help with isolation with regards to operating
+system development. 1) Supporting trait bounds on functions and closures with any number of
+arguments 2) Expose type information in procedural macros 3) Support a collision-free, unique type
+identifier 4) Support extendable, no\_std unwind library 5) Support typed assembly language for Rust
+6) Support trusted build environments 7) Provide software-only stack guard with extensible probing
+interface 8) Develop zero-copy serialization of “plain- old” data structures. The authors argue that
+with these changes or inclusions to the Rust programming language developing operating system
+kernels would be much easier and safer.
+
+Klimt et al.~\cite{Klimt2023-ob} details the lessons learned in implementing Theseus, an operating
+system written in Rust. The authors describe how Theseus uses intralingual design to maximize the
+compilers role in enforcing correctness. By leveraging Rusts type system and borrow checker memory
+safety and correct ownership transfer can be achieved at a higher level than what could be done in
+C. The authors also detailed some of the limitations of intralingual design such as not being as
+expressive as many other formal verification techniques due to the limited invariant that can be
+enforced by the type system. One of the most important lessons learned was the insight that a linear
+type system itself cannot guarantee uniqueness of the resource represented, such when a memory
+resources may overlap. The authors introduce the idea of using a hybrid approach of verification
+where they leverage both the linear type system and an SMT solver.
+
+Klimt et al.~\cite{Klimt2023-ob} explore the challenges of using rust to write an operating
+system. It is impossible to write a complete operating system in 100\% safe rust. For example when
+writing a memory management system raw pointers must be used in order to modify the hardware. The
+authors found four typical soundness issues in their code base. 1) Unsynchronized Global State - any
+use of mutable statics is unsafe, 2) C-Style Abstractions - Use rust style abstractions to properly
+encapsulate interal unsafe usage of raw pointers, 3) Aliasing Mutable References - Giving out raw
+pointers to memory that is also referenced mutably, and 4) Reimplementing Memory Access - instead of
+accessing specific memory regions in assembly, creating and using references to the whole region is
+preferable. The authors also explore the bootstrapping problem with rust systems. How do we ensure
+that ownership of memory is sound in the operating system itself? In operating systems written in C
+the kernel provides the ownership root to applications running on top, finding a new ownership root
+is a open research question with regards to using Rust for operating system development.
 
@@
 \item \href{https://hubris.oxide.computer/}{Hubris OS}
-\item Theseus
+\item Theseus (Mono)
+\item Redleaf
+\item Splinter
+\item Redox
+\item Drone OS
+\item Bern RTOS
+\item Hermit (unikernel)
+  \item Atmoshpere (Micro)
 \end{itemize}

### 2024-05-20 paper.tex
Update on Overleaf.

@@
 duplicates that needed to be removed. We then used Google Scholar for forward and backward
-snowballing.
+snowballing. We choose to look at only the previous 5 years of research in order to capture the bleeding edge research that is currently being done.
 
@@
 
-\subsection{RQ1}
+\subsection{RQ1 What are the existing approaches and methodologies for implementing
+      operating system kernels in Rust?}
+
+Table~\ref{tab:RQ1} summarizes our findings regarding what existing approaches researchers are using to integrate rust into the kernel.
 
@@
 programming language in order to take advantage of Rust's memory safe features. The new Unikernel
-consists of only 3.27\% unsafe Rust with the rest of the code base consisting of safe Rust
+consists of only 3.27\% unsafe Rust with the rest of the code base consisting of safe Rust.
 
-\subsection{RQ2}
+\begin{table*}
+    \begin{tabular}{||p{2cm}|p{2cm}|p{8cm}||}
+    \hline
+    Approach & Papers & Operating System\\
+    \hline\hline
+    Monolithic  & ~\cite{The_kernel_development_community_undated-iw} & \href{https://docs.kernel.org/rust/}{Linux kernel v6.1+}\\
+    Micro-kernel & ~\cite{Chen2023-wb} & Atmosphere, Redox, Redleaf\\
+    Embedded & ~\cite{Culic2022-bk} & \href{https://github.com/tock/tock}{Tock} \\
+    Unikernel & ~\cite{Lankes2019-cm} & RustyHermit \\
+    Prototype & ~\cite{Boos2020-zh, Ijaz2023-da} & Thesus \\
+    \hline
+  \end{tabular}
+  \caption{Approaches and Methodologies for Rust in the Kernel}
+    \label{tab:RQ1}
+\end{table*}
+
+This is a list of operating systems.
+\begin{itemize}
+\item 
+\item 
+\item \href{https://hubris.oxide.computer/}{Hubris OS}
+\item Theseus (Mono)
+
+\item Splinter
+
+\item Drone OS
+\item Bern RTOS
+
+\end{itemize}
+
+\subsection{RQ2  What are the performance implications of using Rust for operating system
+      kernel development in terms of throughput, latency, and resource utilization?}
 
@@
 
-\subsection{RQ3}
+\subsection{RQ3 What are the major challenges, limitations, and lessons learned when
+      developing operating system kernels in Rust, particularly in comparison to other languages?}
 
@@
 \begin{itemize}
-
     \item Deeply ingrained monomorphization
@@
 
-\begin{table}
-  \begin{tabular}{||c|c||}
-    \hline
-    System Type & Papers\\
-    \hline\hline
-    Embedded & ~\cite{Ayers2022-vk} \\
-    Non-Embedded & 0 \\
-    Unikernel & 0 \\
-    Kernel & 0 \\
-    \hline
-  \end{tabular}
-\end{table}
-
-This is a list of operating systems.
-\begin{itemize}
-\item \href{https://docs.kernel.org/rust/}{Linux kernel v6.1+}
-\item \href{https://github.com/tock/tock}{Tock}
-\item \href{https://hubris.oxide.computer/}{Hubris OS}
-\item Theseus (Mono)
-\item Redleaf
-\item Splinter
-\item Redox
-\item Drone OS
-\item Bern RTOS
-\item Hermit (unikernel)
-  \item Atmoshpere (Micro)
-\end{itemize}
-
 \section{Conclusion}

### 2024-05-23 paper.tex
Update on Overleaf.

@@
 \usepackage[raggedrightboxes]{ragged2e}
+
 \pagestyle{plain}
@@
 
-The 1995 movie Hackers~\cite{Wikipedia_contributors2024-zr} prescient predictions regarding the ease
-of breaking into computing systems in cyberspace have came to fruition. The White House Office of
-the National Cyber Director (ONCD) released a report calling for the technical community to
-proactively reduce the attack surface in cyberspace with a two pronged approach. First we need to
-address the root cause of many of the most heinous cyber attacks, memory unsafe programming
-languages~\cite{United_States_Gov2024-pp}. Second, we need to establish better cybersecurity quality
-metrics so we can have a better understanding of the cyber security landscape.
+The 1995 movie Hackers~\cite{Wikipedia_contributors2024-zr} prescient predictions regarding the ease of breaking into computing systems in cyberspace have came to fruition.
+The White House Office of the National Cyber Director (ONCD) released a report calling for the technical community to proactively reduce the attack surface in cyberspace with a two pronged approach.
+First we need to address the root cause of many of the most heinous cyber attacks, memory unsafe programming languages~\cite{United_States_Gov2024-pp}.
+Second, we need to establish better cybersecurity quality metrics so we can have a better understanding of the cyber security landscape.
 
-In the ever-evolving landscape of software development, the reliability and security of computer
-systems stands as a paramount concern for all parties involved. Modern software is constructed by
-building ever more complex abstractions, one on top of the other. Thus, if we aim to have a secure
-system we must start to peel back all the layers and tackle one of the fundamental abstractions in
-computer science, the programming language. Programming languages that provide and enforce memory
-safety eliminate whole classes of bugs such as buffer overflows, dangling pointers, and memory leaks
-which have been implicated in a myriad of security vulnerabilities and system crashes.
+In the ever-evolving landscape of software development, the reliability and security of computer systems stands as a paramount concern for all parties involved.
+Modern software is constructed by building ever more complex abstractions, one on top of the other.
+Thus, if we aim to have a secure system we must start to peel back all the layers and tackle one of the fundamental abstractions in computer science, the programming language.
+Programming languages that provide and enforce memory safety eliminate whole classes of bugs such as buffer overflows, dangling pointers, and memory leaks which have been implicated in a myriad of security vulnerabilities and system crashes.
 
-The migration of legacy C codebases to memory-safe programming languages like Rust is of paramount
-importance in modern software development, aiming to mitigate common security vulnerabilities and
-memory-related errors. This paper presents a systematic literature review (SLR) focusing on
-strategies and methodologies for integrating Rust into one of the most fundamental areas that are
-typically dominated by unsafe languages the operating system kernel. Our study aims to provide a
-comprehensive overview of existing research, identify gaps, and suggest future directions in this
-domain. Through a rigorous search process, we synthesized relevant studies and extracted key
-findings to offer insights into effective approaches for ensuring memory safety when working closley
-with hardware.
+The migration of legacy C code bases to memory-safe programming languages like Rust is of paramount importance in modern software development, aiming to mitigate common security vulnerabilities and memory-related errors.
+This paper presents a systematic literature review (SLR) focusing on strategies and methodologies for integrating Rust into one of the most fundamental areas that are typically dominated by unsafe languages the operating system kernel.
+Our study aims to provide a comprehensive overview of existing research, identify gaps, and suggest future directions in this domain.
+Through a rigorous search process, we synthesized relevant studies and extracted key findings to offer insights into effective approaches for ensuring memory safety when working closely with hardware.
 
@@
 
-For our research methodolgy we followed the advice of Kitchenham and Charters~\cite{Stuart2007-cc}
-and divided our review into the 3 discrete phases, planning the review, conducting the review, and
-reporting the review results. The following sections detail our review process.
+For our research methodolgy we followed the advice of Kitchenham and Charters~\cite{Stuart2007-cc} and divided our review into the 3 discrete phases, planning the review, conducting the review, and reporting the review results.
+The following sections detail our review process.
 
@@
 
-Before we start researching we must first confirm the need for a SLR. The recent report released by
-ONCD~\cite{United_States_Gov2024-pp} has conveniently done this job for us by compiling a report
-detailing the need for research in the domain of memory safety. While the ONCD report detailed a two
-pronged approach, for this SLR we will be focusing on a memory safe programming language,
-specifically Rust. While there are many modern memory safe programming languages available for software
-developers to use Rust is one of the few languages that is feasible to use when developing operating
-system kernels due to is lack of runtime and garbage collector. Even the Linux kernel is in the
-early stages of adding real support for Rust~\cite{The_kernel_development_community_undated-iw}
-thus, we will focus on Rust as a primary candidate to replace the aging C programming language.
+Before we start researching we must first confirm the need for a SLR.
+The recent report released by ONCD~\cite{United_States_Gov2024-pp} has conveniently done this job for us by compiling a report detailing the need for research in the domain of memory safety.
+While the ONCD report detailed a two pronged approach, for this SLR we will be focusing on a memory safe programming language, specifically Rust.
+While there are many modern memory safe programming languages available for software developers to use Rust is one of the few languages that is feasible to use when developing operating system kernels due to is lack of runtime and garbage collector.
+Even the Linux kernel is in the early stages of adding real support for Rust~\cite{The_kernel_development_community_undated-iw} thus, we will focus on Rust as a primary candidate to replace the aging C programming language.
 
@@
 
-We employed a robust multi-step search strategy across three digital databases, ACM Digital Library,
-IEEE Xplore, and Scopus in order to find all the current research that has been done. These
-databases cover a wide range of current software engineering and computer science literature. We
-leveraged the advanced search features of all three databases to search the titles and abstracts for
-keywords using boolean search operators.
+We employed a robust multi-step search strategy across three digital databases, ACM Digital Library, IEEE Xplore, and Scopus in order to find all the current research that has been done.
+These databases cover a wide range of current software engineering and computer science literature.
+We leveraged the advanced search features of all three databases to search the titles and abstracts for keywords using boolean search operators.
 
@@
 
-We search all three databases with the keywords listed in table~\ref{tab:keywords} between January
-1, 2019 and April 1, 2024. While both ACM and IEEE had mutually exclusive results Scopus had 44
-duplicates that needed to be removed. We then used Google Scholar for forward and backward
-snowballing. We choose to look at only the previous 5 years of research in order to capture the bleeding edge research that is currently being done.
+We search all three databases with the keywords listed in table~\ref{tab:keywords} between January 1, 2019 and April 1, 2024.
+While both ACM and IEEE had mutually exclusive results Scopus had 44 duplicates that needed to be removed.
+We then used Google Scholar for forward and backward snowballing.
+We choose to look at only the previous 5 years of research in order to capture the bleeding edge research that is currently being done.
 
@@
 
-Following the guidelines outlined by Kitchenham and Charters~\cite{Stuart2007-cc} we set the
-inclusion and exclusion criteria based on our research questions outlined in
-section~\ref{sec:researchQuestions}. We only consider papers that are written in English, published
-in conferences or journals, and within the time frame of 2019 and 2024. The published papers should
-describe using the Rust programming language for either developing a new kernel, extending an
-existing kernel, or authoring drivers. We will include any type of kernel architecture including
-monolithic kernel, microkernel, or unikernel in both the embedded and non-embedded space as long as
-the paper is using the Rust programming language in some way. Papers that describe solutions that
-reside 100\% in user space will be excluded.
+Following the guidelines outlined by Kitchenham and Charters~\cite{Stuart2007-cc} we set the inclusion and exclusion criteria based on our research questions outlined in section~\ref{sec:researchQuestions}.
+We only consider papers that are written in English, published in conferences or journals, and within the time frame of 2019 and 2024.
+The published papers should describe using the Rust programming language for either developing a new kernel, extending an existing kernel, or authoring drivers.
+We will include any type of kernel architecture including monolithic kernel, microkernel, or unikernel in both the embedded and non-embedded space as long as the paper is using the Rust programming language in some way.
+Papers that describe solutions that reside 100\% in user space will be excluded.
 
@@
 
-Once we had our initial collection of papers from the databases we merged all duplicate
-papers into one record and then started our first pass which consisted of reading the title and
-abstract and either marking the paper as \textit{include} or \textit{exclude}.
-Once  the initial screening was completed we were left with 32 potential papers that
-we needed an in depth reading.
+Once we had our initial collection of papers from the databases we merged all duplicate papers into one record and then started our first pass which consisted of reading the title and abstract and either marking the paper as \textit{include} or \textit{exclude}.
+Once  the initial screening was completed we were left with 29 potential papers that we needed an in depth reading.
 
@@
 
-In this section we present our findings. We address each research question from RQ1 to RQ3.
+In this section we present our findings. We address each research question from \textbf{RQ1} - \textbf{RQ3}.
 
-\subsection{RQ1 What are the existing approaches and methodologies for implementing
-      operating system kernels in Rust?}
+\subsection{RQ1 What are the existing approaches and methodologies for implementing operating system kernels in Rust?}
 
@@
 
-Boos et al.~\cite{Boos2020-zh} is about an experimental operating system that is aimed at addressing
-the state spill problem. State spill happens when a single service in an operating system can harbor
-a state change, induced by interacting with other services, that can eventually cause a system crash
-or system instability. The operating system is built in Rust and leverages the properties of Rust to
-construct a safer system. Theseus operates in a single address space and single privilege level and
-uses properties of the Rust programming language to realize isolation instead of relying on
-hardware. The design of the OS uses a novel cell based structure where ownership of memory and
-resources is enforced by the compiler. In short the Theseus OS is an operating system that was
-designed to match the Rust language instead of the more traditional route of matching the hardware.
+Boos et al.~\cite{Boos2020-zh} is about an experimental operating system that is aimed at addressing the state spill problem.
+State spill happens when a single service in an operating system can harbor a state change, induced by interacting with other services, that can eventually cause a system crash or system instability.
+The operating system is built in Rust and leverages the properties of Rust to construct a safer system.
+Theseus operates in a single address space and single privilege level and uses properties of the Rust programming language to realize isolation instead of relying on hardware.
+The design of the OS uses a novel cell based structure where ownership of memory and resources is enforced by the compiler.
+In short the Theseus OS is an operating system that was designed to match the Rust language instead of the more traditional route of matching the hardware.
 
-Chen X et al.~\cite{Chen2023-wb} primary focused on developing a micro-kernel in Rust that could be
-formally verified by using both the liner type system or rust and a SMT solver. The new micro-kernel
-named Atmosphere follows the tradition of other micro-kernels by pushing most kernel functionally to
-users-space thus limiting the surface area that needs to be proved. The authors were able to get a
-7.5:1 proof-to-code ratio which is higher that other formerly verified microkernels SeL4 and
-CeriKOS, which both have proof-tocode ratio of 19:1 and 20:1 respectively.
+Chen X et al.~\cite{Chen2023-wb} primary focused on developing a micro-kernel in Rust that could be formally verified by using both the liner type system or rust and a SMT solver.
+The new micro-kernel named Atmosphere follows the tradition of other micro-kernels by pushing most kernel functionally to users-space thus limiting the surface area that needs to be proved.
+The authors were able to get a 7.5:1 proof-to-code ratio which is higher that other formerly verified microkernels SeL4 and CeriKOS, which both have proof to code ratio of 19:1 and 20:1 respectively.
 
-Lankes et al.~\cite{Lankes2019-cm} detailed their experiences with developing a Unikernel in
-Rust. Originally written in C and called HermitCore the authors ported the Kernel to the Rust
-programming language in order to take advantage of Rust's memory safe features. The new Unikernel
-consists of only 3.27\% unsafe Rust with the rest of the code base consisting of safe Rust.
+Lankes et al.~\cite{Lankes2019-cm} detailed their experiences with developing a Unikernel in Rust.
+Originally written in C and called HermitCore the authors ported the Kernel to the Rust programming language in order to take advantage of Rust's memory safe features.
+The new Unikernel consists of only 3.27\% unsafe Rust with the rest of the code base consisting of safe Rust.
 
-\begin{table*}
-    \begin{tabular}{||p{2cm}|p{2cm}|p{8cm}||}
+Li and Sato~\cite{Li2024-yb} explored using Rust to write an exokernel named W-Kernel.
+The authors proposed a novel architecture for an OS by embedding a WebAssembly(WASM) runtime into the kernel what can run programs written in any language that can compile down to WASM.
+
+Liang et al.~\cite{Liang2021-bo} developed a new operating system named Rustpi with a micro-kernel architecture implemented in Rust to explore how modern language features can help with building reilabe systems.
+
+\begin{table}
+    \begin{tabular}{||l|l|l||}
     \hline
-    Approach & Papers & Operating System\\
+    Approach & Papers & Operating System in Rust\\
     \hline\hline
-    Monolithic  & ~\cite{The_kernel_development_community_undated-iw} & \href{https://docs.kernel.org/rust/}{Linux kernel v6.1+}\\
-    Micro-kernel & ~\cite{Chen2023-wb} & Atmosphere, Redox, Redleaf\\
-    Embedded & ~\cite{Culic2022-bk} & \href{https://github.com/tock/tock}{Tock} \\
-    Unikernel & ~\cite{Lankes2019-cm} & RustyHermit \\
-    Prototype & ~\cite{Boos2020-zh, Ijaz2023-da} & Thesus \\
+    Monolithic  & \cite{The_kernel_development_community_undated-iw}, \cite{Li2019-ru} & \href{https://docs.kernel.org/rust/}{Linux kernel v6.1+}\\
+    Micro-kernel & \cite{Chen2023-wb},\cite{Liang2021-bo}, \cite{Liu2024-xe} & Atmosphere, Redox, Redleaf\\
+    Embedded & \cite{Culic2022-bk} & \href{https://github.com/tock/tock}{Tock}, \href{https://hubris.oxide.computer/}{Hubris}, \href{https://www.drone-os.com/}{Drone}, \href{https://bern-rtos.org/}{Bern} \\
+    Unikernel & \cite{Lankes2019-cm},  \cite{Boos2020-zh}, \cite{Ijaz2023-da}  & RustyHermit, Theseus \\
+    Exokernel & \cite{Li2024-yb} & W-Kernel \\
     \hline
@@
     \label{tab:RQ1}
-\end{table*}
-
-This is a list of operating systems.
-\begin{itemize}
-\item 
-\item 
-\item \href{https://hubris.oxide.computer/}{Hubris OS}
-\item Theseus (Mono)
-
-\item Splinter
-
-\item Drone OS
-\item Bern RTOS
-
-\end{itemize}
+\end{table}
 
@@
       kernel development in terms of throughput, latency, and resource utilization?}
+
+Comparing performance, throughput, latency, and resource utilization between different systems is very difficult.
+For example several of the operating systems written in Rust such as Tock are a complete rethinking of how an operating system is designed and Rust for Linux (RFL) is still relatively new and has not had the benefit of decades of incremental performance improvements like components that are written in C.
+Additionally, direct comparisons between a Monolithic kernel, Micro-kernel, and Unikernel, or comparisons between a Real Time Operation System (RTOS) and a General Purpose operating System are not directly meaningful due to the vastly different goals of each system.
+Therefore, we will focus more on the overall efforts that are specific to the Rust programming language and the challenges presented.
+We summarize our findings in table~\ref{tab:RQ2}.
 
@@
 
-Li et al.~\cite{Li2019-ru} explored the feasibility of using Rust in kernel space. The authors took
+Li et al.~\cite{Li2024-be} explored the feasibility of using Rust in kernel space. The authors took
 an existing component, the Out of Memory (OOM) and implemented a replacement using the Rust
@@
 
+\begin{table}
+    \begin{tabular}{||l|l|l||}
+    \hline
+    No. & Implication & Studies that Reported the challenge\\
+    \hline\hline
+    1 & Performance & \cite{Gonzalez2023-ek}, \cite{Li2024-be}, \cite{Ma2023-ef}\\
+    2 & Throughput & \cite{Gonzalez2023-ek}\\
+    3 & Latency & \cite{Culic2022-bk} \\
+    4 & Resource Utilization & 0 \\
+    \hline
+  \end{tabular}
+  \caption{Performance Implications of using Rust in the Kernel}
+    \label{tab:RQ2}
+\end{table}
+
 \subsection{RQ3 What are the major challenges, limitations, and lessons learned when
@@
 
-Ayers et al.~\cite{Ayers2022-sf} focused on the size of rust binaries in the embedded
-environment. Their work primary focused on the size of binaries produced using the embedded OS
-Tock. The authors identified four major causes of binary growth:
+\begin{table*}[htbp]
+    \begin{tabular}{p{1cm} p{5cm} p{10cm}}
+        \hline
+        No. & Challenge & Description\\
+        \hline
+        1 & Binary Size~\cite{Ayers2022-sf}  & 
+        \begin{itemize}
+            \item Deeply ingrained monomorphization which increased the size of the rust binaries
+            \item Compiler optimizations are not as mature as some C based compilers thus increasing binary size
+            \item Hidden data structures and data
+            \item Sub-optimal compiler generated support code
+        \end{itemize}
+        \\
+        \hline
+        2 & Missing Features~\cite{Burtsev2021-mh} &
+        \begin{itemize}
+            \item Supporting trait bounds on functions and closures with any number of arguments
+            \item Expose type information in procedural macros
+            \item Support a collision-free, unique type identifier
 
+            \item Support typed assembly language for Rust
+            \item Support trusted build environments 
+            \item Provide software-only stack guard with extensible probing interface
+            \item Develop zero-copy serialization of “plain- old” data structures
+        \end{itemize}
+        \\
+        \hline
+        3 & Soundness~\cite{Klimt2023-ob} &
         \begin{itemize}
-    \item Deeply ingrained monomorphization
-    \item Suboptimal compiler generated support code
-    \item Hidden data structures and data
-    \item Fewer compiler optimizations
+            \item Unsynchronized Global State - any use of mutable statics is unsafe
+            \item C-Style Abstractions - Use rust style abstractions to properly encapsulate internal unsafe usage of raw pointers
+            \item Aliasing Mutable References - Giving out raw pointers to memory that is also referenced mutably
+            \item Re-implementing Memory Access - instead of accessing specific memory regions in assembly, creating and using references to the whole region is preferable
         \end{itemize}
+        \\
+        \hline
+        4 & Panics \cite{Ma2023-ef},\cite{Burtsev2021-mh} &
+        \begin{itemize}
+            \item Support extendable, no\_std unwind library
+            \item Stack unwinding in embedded environments
+        \end{itemize}
+        \\
+    \hline
+  \end{tabular}
+  \caption{Challenges Unique to the Rust Programming Language}
+    \label{tab:RQ3}
+\end{table*}
 
-In addition to the identified causes of binary growth the Ayers et al.~\cite{Ayers2022-sf} have the
-following 5 recommendations when using Rust in a size constrained environment:
+Ayers et al.~\cite{Ayers2022-sf} focused on the size of rust binaries in the embedded environment.
+Their work primary focused on the size of binaries produced using the embedded OS Tock.
+In addition to the identified causes of binary growth the authors have the following 5 recommendations when using Rust in a size constrained environment:
 \begin{itemize}
@@
 
-Burtsev et al.~\cite{Burtsev2021-mh} explores what is missing in the Rust programming language to
-help solve the isolation problem in an operating system. Currently Rust lacks the ability to express
-isolation in the heap without external support. The RedLeaf experiential operating system relied on
-an complex interface definition language (IDL) to enforce isolated heaps. The paper enumerates
-several properties of the Rust Language that could help with isolation with regards to operating
-system development. 1) Supporting trait bounds on functions and closures with any number of
-arguments 2) Expose type information in procedural macros 3) Support a collision-free, unique type
-identifier 4) Support extendable, no\_std unwind library 5) Support typed assembly language for Rust
-6) Support trusted build environments 7) Provide software-only stack guard with extensible probing
-interface 8) Develop zero-copy serialization of “plain- old” data structures. The authors argue that
-with these changes or inclusions to the Rust programming language developing operating system
-kernels would be much easier and safer.
+Burtsev et al.~\cite{Burtsev2021-mh} explores what is missing in the Rust programming language to help solve the isolation problem in an operating system.
+Currently Rust lacks the ability to express isolation in the heap without external support.
+The RedLeaf experiential operating system relied on an complex interface definition language (IDL) to enforce isolated heaps.
+The paper enumerates several properties of the Rust Language that could help with isolation with regards to operating system development that are detailed in table~\ref{tab:RQ3}.
+The authors argue that with these changes or inclusions to the Rust programming language developing operating system kernels would be much easier and safer.
 
-Klimt et al.~\cite{Klimt2023-ob} details the lessons learned in implementing Theseus, an operating
-system written in Rust. The authors describe how Theseus uses intralingual design to maximize the
-compilers role in enforcing correctness. By leveraging Rusts type system and borrow checker memory
-safety and correct ownership transfer can be achieved at a higher level than what could be done in
-C. The authors also detailed some of the limitations of intralingual design such as not being as
-expressive as many other formal verification techniques due to the limited invariant that can be
-enforced by the type system. One of the most important lessons learned was the insight that a linear
-type system itself cannot guarantee uniqueness of the resource represented, such when a memory
-resources may overlap. The authors introduce the idea of using a hybrid approach of verification
-where they leverage both the linear type system and an SMT solver.
-
-Klimt et al.~\cite{Klimt2023-ob} explore the challenges of using rust to write an operating
-system. It is impossible to write a complete operating system in 100\% safe rust. For example when
-writing a memory management system raw pointers must be used in order to modify the hardware. The
-authors found four typical soundness issues in their code base. 1) Unsynchronized Global State - any
-use of mutable statics is unsafe, 2) C-Style Abstractions - Use rust style abstractions to properly
-encapsulate interal unsafe usage of raw pointers, 3) Aliasing Mutable References - Giving out raw
-pointers to memory that is also referenced mutably, and 4) Reimplementing Memory Access - instead of
-accessing specific memory regions in assembly, creating and using references to the whole region is
-preferable. The authors also explore the bootstrapping problem with rust systems. How do we ensure
-that ownership of memory is sound in the operating system itself? In operating systems written in C
-the kernel provides the ownership root to applications running on top, finding a new ownership root
-is a open research question with regards to using Rust for operating system development.
+Klimt et al.~\cite{Klimt2023-ob} details the lessons learned and challenges when implementing Theseus, an operating system written in Rust.
+First they found that it is impossible to write a complete operating system in 100\% safe rust.
+For example when writing a memory management system raw pointers must be used in order to modify the hardware.
+Despite the limitations imposed by the hardware the authors describe how Theseus leveraged intralingual design to maximize the compilers role in enforcing correctness.
+By leveraging Rusts type system and borrow checker memory safety and correct ownership transfer can be achieved at a higher level than what could be done in C.
+The authors also detailed some of the limitations of intralingual design such as not being as expressive as many other formal verification techniques due to the limited invariant that can be enforced by the type system.
+One of the most important lessons learned was the insight that a linear type system itself cannot guarantee uniqueness of the resource represented, such when a memory resources may overlap.
+The authors introduce the idea of using a hybrid approach of verification where they leverage both the linear type system and an SMT solver.
+The authors also explore the bootstrapping problem with rust systems.
+How do we ensure that ownership of memory is sound in the operating system itself?
+In operating systems written in C the kernel provides the ownership root to applications running on top, finding a new ownership root is a open research question with regards to using Rust for operating system development.
 

### 2024-05-24 paper.tex
Update on Overleaf.

@@
 %%
+
 \begin{CCSXML}
@@
 <concept>
-  <concept_id>00000000.0000000.0000000</concept_id>
-  <concept_desc>Do Not Use This Code, Generate the Correct Terms for Your Paper</concept_desc>
+<concept_id>10002944.10011122.10002945</concept_id>
+<concept_desc>General and reference~Surveys and overviews</concept_desc>
 <concept_significance>500</concept_significance>
@@
 <concept>
-  <concept_id>00000000.00000000.00000000</concept_id>
-  <concept_desc>Do Not Use This Code, Generate the Correct Terms for Your Paper</concept_desc>
-  <concept_significance>300</concept_significance>
+<concept_id>10011007.10011006.10011008.10011024</concept_id>
+<concept_desc>Software and its engineering~Language features</concept_desc>
+<concept_significance>500</concept_significance>
 </concept>
 <concept>
-  <concept_id>00000000.00000000.00000000</concept_id>
-  <concept_desc>Do Not Use This Code, Generate the Correct Terms for Your Paper</concept_desc>
-  <concept_significance>100</concept_significance>
- </concept>
- <concept>
-  <concept_id>00000000.00000000.00000000</concept_id>
-  <concept_desc>Do Not Use This Code, Generate the Correct Terms for Your Paper</concept_desc>
-  <concept_significance>100</concept_significance>
+<concept_id>10010520.10010521</concept_id>
+<concept_desc>Computer systems organization~Architectures</concept_desc>
+<concept_significance>500</concept_significance>
 </concept>
@@
 
-\ccsdesc[500]{Do Not Use This Code~Generate the Correct Terms for Your Paper}
-\ccsdesc[300]{Do Not Use This Code~Generate the Correct Terms for Your Paper}
-\ccsdesc{Do Not Use This Code~Generate the Correct Terms for Your Paper}
-\ccsdesc[100]{Do Not Use This Code~Generate the Correct Terms for Your Paper}
+\ccsdesc[500]{General and reference~Surveys and overviews}
+\ccsdesc[500]{Software and its engineering~Language features}
+\ccsdesc[500]{Computer systems organization~Architectures}
 
@@
 %% the work being presented. Separate the keywords with commas.
-\keywords{Memory safety}
+\keywords{Memory safety, Rust, Kernel, Operating System}
 
-\received{20 February 2007}
-\received[revised]{12 March 2009}
-\received[accepted]{5 June 2009}
+\received{30 May 2024}
+\received[revised]{TBD}
+\received[accepted]{TBD}
 
@@
 Once we had our initial collection of papers from the databases we merged all duplicate papers into one record and then started our first pass which consisted of reading the title and abstract and either marking the paper as \textit{include} or \textit{exclude}.
-Once  the initial screening was completed we were left with 29 potential papers that we needed an in depth reading.
+Once  the initial screening was completed we were left with 28 potential papers that we needed an in depth reading.
 
@@
 
-Liang et al.~\cite{Liang2021-bo} developed a new operating system named Rustpi with a micro-kernel architecture implemented in Rust to explore how modern language features can help with building reilabe systems.
+Liang et al.~\cite{Liang2021-bo} developed a new operating system named Rustpi with a micro-kernel architecture implemented in Rust to explore how modern language features can help with building reliable systems.
 
@@
     \hline\hline
-    Monolithic  & \cite{The_kernel_development_community_undated-iw}, \cite{Li2019-ru} & \href{https://docs.kernel.org/rust/}{Linux kernel v6.1+}\\
-    Micro-kernel & \cite{Chen2023-wb},\cite{Liang2021-bo}, \cite{Liu2024-xe} & Atmosphere, Redox, Redleaf\\
+    Monolithic  & \cite{The_kernel_development_community_undated-iw}, \cite{Li2019-ru}, \cite{Miller2021-pg} & \href{https://docs.kernel.org/rust/}{Linux kernel v6.1+}\\
+    Micro-kernel & \cite{Chen2023-wb},\cite{Liang2021-bo}, \cite{Liu2024-xe} \cite{} & Atmosphere, Redox, Redleaf\\
     Embedded & \cite{Culic2022-bk} & \href{https://github.com/tock/tock}{Tock}, \href{https://hubris.oxide.computer/}{Hubris}, \href{https://www.drone-os.com/}{Drone}, \href{https://bern-rtos.org/}{Bern} \\
@@
 
-Culic et al.~\cite{Culic2022-bk} looked at latency issues in the Tock OS. Tock is a new operating
-system that is designed to run on embedded systems but does not provide Real time capabilities. The
-authors attempted to add real time capabilities by integrating eBPF into the Tock kernel to improve
-the interrupt handlers response time. The Authors found that early work (still in the prototype
-stage) lowers the response times of the system and makes interrupt response times 3x.
+Culic et al.~\cite{Culic2022-bk} looked at latency issues in the Tock OS.
+Tock is a new operating system that is designed to run on embedded systems but does not provide Real time capabilities.
+The authors attempted to add real time capabilities by integrating eBPF into the Tock kernel to improve the interrupt handlers response time.
+The Authors found that early work (still in the prototype stage) lowers the response times of the system and makes interrupt response times 3x.
 
-Gonzalez et al.~\cite{Gonzalez2023-ek} explored using the Rust for Linux Project to implement a
-native UDP driver in Rust in order to explore the performance the Rust programming language. The
-authors were able to get a basic driver working with performance only slightly slower than C using
-the Rust for Linux (RFL) project. The RFL project is still to immature to get a full driver up and
-running but is at a stage were we researchers can start experimenting with different
-approaches.
+Gonzalez et al.~\cite{Gonzalez2023-ek} explored using the Rust for Linux Project to implement a native UDP driver in Rust in order to explore the performance the Rust programming language.
+The authors were able to get a basic driver working with performance only slightly slower than C using the Rust for Linux (RFL) project.
+The RFL project is still to immature to get a full driver up and running but is at a stage were we researchers can start experimenting with different approaches.
 
-Li et al.~\cite{Li2024-be} explored the feasibility of using Rust in kernel space. The authors took
-an existing component, the Out of Memory (OOM) and implemented a replacement using the Rust
-programming language. The non-encapsulated interface Rust component which was almost identical to
-the original C component only introduced a 0.7\% overhead. The encapsulated Rust component on the
-other hand added a 3\% performance overhead. Another important aspect of the kernel is size, we must
-be careful to keep the size of the Rust component as close as we can to the C component so we can
-still run on all the same hardware. The authors found that there was only a 0.06\% size increase
-when compared to the original C implementation.
+Li et al.~\cite{Li2024-be} explored the feasibility of using Rust in kernel space.
+The authors took an existing component, the Out of Memory (OOM) and implemented a replacement using the Rust programming language.
+The non-encapsulated interface Rust component which was almost identical to the original C component only introduced a 0.7\% overhead.
+The encapsulated Rust component on the other hand added a 3\% performance overhead.
+Another important aspect of the kernel is size, we must be careful to keep the size of the Rust component as close as we can to the C component so we can still run on all the same hardware.
+The authors found that there was only a 0.06\% size increase when compared to the original C implementation.
 
@@
         \hline
+        5 & C Interop \cite{Miller2019-xm} &
+        \begin{itemize}
+            \item Kernel interfaces, while designed for extensibility, are not designed for type safety
+            \item Hybrid Code Flow. The rust compiler can not track ownership when switching between modules written in C and Rust
+        \end{itemize}
+        \\
+    \hline
   \end{tabular}

### 2024-05-24 paper.tex
Update on Overleaf.

@@
 %%
-
 \begin{CCSXML}
@@
 Once we had our initial collection of papers from the databases we merged all duplicate papers into one record and then started our first pass which consisted of reading the title and abstract and either marking the paper as \textit{include} or \textit{exclude}.
-Once  the initial screening was completed we were left with 28 potential papers that we needed an in depth reading.
+Once  the initial screening was completed we were left with 26 potential papers that we needed an in depth reading.
 
@@
     \hline\hline
-    Monolithic  & \cite{The_kernel_development_community_undated-iw}, \cite{Li2019-ru}, \cite{Miller2021-pg} & \href{https://docs.kernel.org/rust/}{Linux kernel v6.1+}\\
-    Micro-kernel & \cite{Chen2023-wb},\cite{Liang2021-bo}, \cite{Liu2024-xe} \cite{} & Atmosphere, Redox, Redleaf\\
-    Embedded & \cite{Culic2022-bk} & \href{https://github.com/tock/tock}{Tock}, \href{https://hubris.oxide.computer/}{Hubris}, \href{https://www.drone-os.com/}{Drone}, \href{https://bern-rtos.org/}{Bern} \\
-    Unikernel & \cite{Lankes2019-cm},  \cite{Boos2020-zh}, \cite{Ijaz2023-da}  & RustyHermit, Theseus \\
+    Monolithic  & \cite{The_kernel_development_community_undated-iw}, \cite{Li2019-ru}, \cite{Miller2021-pg}, \cite{Oikawa2023-ms} & \href{https://docs.kernel.org/rust/}{Linux kernel v6.1+}\\
+    Micro-kernel & \cite{Chen2023-wb},\cite{Liang2021-bo}, \cite{Liu2024-xe}, \cite{Narayanan2020-gs}, \cite{Narayanan2019-fd} & Atmosphere, Redox, Redleaf\\
+    Embedded & \cite{Culic2022-bk}, \cite{Vishnunaryan2022-yd} & \href{https://github.com/tock/tock}{Tock}, \href{https://hubris.oxide.computer/}{Hubris}, \href{https://www.drone-os.com/}{Drone}, \href{https://bern-rtos.org/}{Bern}, HarSaRK \\
+    Unikernel & \cite{Lankes2019-cm},  \cite{Boos2020-zh}, \cite{Ijaz2023-da}, \cite{Sung2020-bb}  & RustyHermit, Theseus \\
     Exokernel & \cite{Li2024-yb} & W-Kernel \\

### 2024-05-28 paper.tex
Update on Overleaf.
Updates from the holiday weekend

@@
 \title{Kernel Development Using Rust: A Systematic Literature Review}
-%\title{}
+
+\author{Anonymous Authors}
 
@@
 
-\renewcommand{\shortauthors}{Panter et al.}
+%\renewcommand{\shortauthors}{Panter et al.}
 
 \begin{abstract}
-    \textit{Context:}
-    TODO:
-    \textit{Objective:}
-    TODO:
-    \textit{Method:}
-    TODO:
-    \textit{Results:}
-    TODO:
-    \textit{Conclusions:}
+
+  \textit{Context:} The integration of Rust into kernel development is a transformative endeavor aimed
+  at enhancing system security and reliability by leveraging Rust's strong memory safety
+  guarantees. This systematic literature review examines the current state of research and practice
+  regarding the use of Rust in the kernel development.
+  \textit{Objective:} We aim to find the current edge of research regarding using Rust in Kernel
+  development so we can push the research forward and reduce the number of memory safety
+  vulnerabilities in one of the most critical pieces of software that underpins all modern
+  applications.
+  \textit{Method:} By analyzing a broad spectrum of studies, we identify the advantages Rust offers,
+  such as reducing vulnerabilities like buffer overflows and use-after-free errors, which are
+  prevalent in C-based kernel development. The review also highlights the challenges faced,
+  including integration hurdles with existing C codebases, performance considerations, and the need
+  for community consensus on Rust's adoption.
+  \textit{Results:} Our findings suggest that while the initial implementations of Rust in the kernel
+  shows promising results in terms of safety and stability, significant challenges remain. These
+  include achieving seamless interoperability with existing kernel components, maintaining
+  performance, and ensuring adequate support and tooling for developers.
+  \textit{Conclusions:} In conclusion, this review provides a comprehensive overview of the
+  potential and challenges of incorporating Rust into kernel development. It underscores the need for
+  continued research and practical implementation efforts to fully realize the benefits of Rust. By
+  addressing these challenges, the integration of Rust could mark a significant step forward in the
+  evolution of operating system development towards safer and more reliable systems.
+
 \end{abstract}
@@
 %% the work being presented. Separate the keywords with commas.
-\keywords{Memory safety, Rust, Kernel, Operating System}
+\keywords{Memory safety, Rust, Kernel, Operating System, Linux}
 
@@
 
-The 1995 movie Hackers~\cite{Wikipedia_contributors2024-zr} prescient predictions regarding the ease of breaking into computing systems in cyberspace have came to fruition.
-The White House Office of the National Cyber Director (ONCD) released a report calling for the technical community to proactively reduce the attack surface in cyberspace with a two pronged approach.
-First we need to address the root cause of many of the most heinous cyber attacks, memory unsafe programming languages~\cite{United_States_Gov2024-pp}.
-Second, we need to establish better cybersecurity quality metrics so we can have a better understanding of the cyber security landscape.
+The 1995 movie Hackers~\cite{Wikipedia_contributors2024-zr} prescient predictions regarding the ease
+of breaking into computing systems in cyberspace have came to fruition. The White House Office of
+the National Cyber Director (ONCD) released a report calling for the technical community to
+proactively reduce the attack surface in cyberspace with a two pronged approach. First we need to
+address the root cause of many of the most heinous cyber attacks, memory unsafe programming
+languages~\cite{United_States_Gov2024-pp}. Second, we need to establish better cybersecurity quality
+metrics so we can have a better understanding of the cyber security landscape.
 
-In the ever-evolving landscape of software development, the reliability and security of computer systems stands as a paramount concern for all parties involved.
-Modern software is constructed by building ever more complex abstractions, one on top of the other.
-Thus, if we aim to have a secure system we must start to peel back all the layers and tackle one of the fundamental abstractions in computer science, the programming language.
-Programming languages that provide and enforce memory safety eliminate whole classes of bugs such as buffer overflows, dangling pointers, and memory leaks which have been implicated in a myriad of security vulnerabilities and system crashes.
+In the ever-evolving landscape of software development, the reliability and security of computer
+systems stands as a paramount concern for all parties involved. Modern software is constructed by
+building ever more complex abstractions, one on top of the other. Thus, if we aim to have a secure
+system we must start to peel back all the layers and tackle one of the fundamental abstractions in
+computer science, the programming language. Programming languages that provide and enforce memory
+safety eliminate whole classes of bugs such as buffer overflows, dangling pointers, and memory leaks
+which have been implicated in a myriad of security vulnerabilities and system crashes.
 
-The migration of legacy C code bases to memory-safe programming languages like Rust is of paramount importance in modern software development, aiming to mitigate common security vulnerabilities and memory-related errors.
-This paper presents a systematic literature review (SLR) focusing on strategies and methodologies for integrating Rust into one of the most fundamental areas that are typically dominated by unsafe languages the operating system kernel.
-Our study aims to provide a comprehensive overview of existing research, identify gaps, and suggest future directions in this domain.
-Through a rigorous search process, we synthesized relevant studies and extracted key findings to offer insights into effective approaches for ensuring memory safety when working closely with hardware.
+Most operating system kernels are predominantly written in the C programming language with bits of
+assembly. C has been favored for its low-level capabilities and performance efficiency, crucial for
+kernel development. However, the inherent lack of memory safety in C has led to numerous security
+vulnerabilities, including buffer overflows and use-after-free errors, which have plagued operation
+system development for decades~\cite{noauthor_undated-bf}. Addressing these vulnerabilities is
+paramount to enhancing the security and reliability of all operating systems.
+
+Rust, a systems programming language developed by Mozilla, has garnered significant attention for
+its strong emphasis on memory safety without sacrificing performance. Rust's ownership model and its
+compile-time checks effectively prevent common programming errors that lead to security
+vulnerabilities. What makes Rust unique is the fact that it accomplishes these tasks without the use
+of a garbage collector, making it an ideal candidate for kernel development where safety and
+performance are critical.
+
+The potential integration of Rust into kernel development represents a significant shift in the
+landscape of operating system development. This shift prompts a comprehensive evaluation of both the
+opportunities and challenges associated with using Rust in such a foundational component of modern
+computing. This systematic literature review aims to synthesize existing research and practical
+experiences related to the use of Rust in the kernel, providing a detailed understanding of
+the current state of this emerging field.
+
+This paper presents a systematic literature review (SLR) focusing on strategies and methodologies
+for integrating Rust into one of the most fundamental areas that are typically dominated by unsafe
+languages the operating system kernel. Our study aims to provide a comprehensive overview of
+existing research, identify gaps, and suggest future directions in this domain. Through a rigorous
+search process, we synthesized relevant studies and extracted key findings to offer insights into
+effective approaches for ensuring memory safety when working closely with hardware.
 
@@
 
-For our research methodolgy we followed the advice of Kitchenham and Charters~\cite{Stuart2007-cc} and divided our review into the 3 discrete phases, planning the review, conducting the review, and reporting the review results.
-The following sections detail our review process.
+For our research methodolgy we followed the advice of Kitchenham and Charters~\cite{Stuart2007-cc}
+and divided our review into the 3 discrete phases, planning the review, conducting the review, and
+reporting the review results. The following sections detail our review process.
 
@@
 
-Before we start researching we must first confirm the need for a SLR.
-The recent report released by ONCD~\cite{United_States_Gov2024-pp} has conveniently done this job for us by compiling a report detailing the need for research in the domain of memory safety.
-While the ONCD report detailed a two pronged approach, for this SLR we will be focusing on a memory safe programming language, specifically Rust.
-While there are many modern memory safe programming languages available for software developers to use Rust is one of the few languages that is feasible to use when developing operating system kernels due to is lack of runtime and garbage collector.
-Even the Linux kernel is in the early stages of adding real support for Rust~\cite{The_kernel_development_community_undated-iw} thus, we will focus on Rust as a primary candidate to replace the aging C programming language.
+Before we start researching we must first confirm the need for a SLR. The recent report released by
+ONCD~\cite{United_States_Gov2024-pp} has conveniently done this job for us by compiling a report
+detailing the need for research in the domain of memory safety. While the ONCD report detailed a two
+pronged approach, for this SLR we will be focusing on a memory safe programming language,
+specifically Rust. While there are many modern memory safe programming languages available for
+software developers to use Rust is one of the few languages that is feasible to use when developing
+operating system kernels due to is lack of runtime and garbage collector. Even the Linux kernel is
+in the early stages of adding real support for
+Rust~\cite{The_kernel_development_community_undated-iw} thus, we will focus on Rust as a primary
+candidate to replace the aging C programming language.
 
@@
 
-We employed a robust multi-step search strategy across three digital databases, ACM Digital Library, IEEE Xplore, and Scopus in order to find all the current research that has been done.
-These databases cover a wide range of current software engineering and computer science literature.
-We leveraged the advanced search features of all three databases to search the titles and abstracts for keywords using boolean search operators.
+We employed a robust multi-step search strategy across three digital databases, ACM Digital Library,
+IEEE Xplore, and Scopus in order to find all the current research that has been done. These
+databases cover a wide range of current software engineering and computer science literature. We
+leveraged the advanced search features of all three databases to search the titles and abstracts for
+keywords using boolean search operators.
 
@@
 
-We search all three databases with the keywords listed in table~\ref{tab:keywords} between January 1, 2019 and April 1, 2024.
-While both ACM and IEEE had mutually exclusive results Scopus had 44 duplicates that needed to be removed.
-We then used Google Scholar for forward and backward snowballing.
-We choose to look at only the previous 5 years of research in order to capture the bleeding edge research that is currently being done.
+We search all three databases with the keywords listed in table~\ref{tab:keywords} between January
+1, 2019 and April 1, 2024. While both ACM and IEEE had mutually exclusive results Scopus had 44
+duplicates that needed to be removed. We then used Google Scholar for forward and backward
+snowballing. We choose to look at only the previous 5 years of research in order to capture the
+bleeding edge research that is currently being done.
 
@@
 
-Following the guidelines outlined by Kitchenham and Charters~\cite{Stuart2007-cc} we set the inclusion and exclusion criteria based on our research questions outlined in section~\ref{sec:researchQuestions}.
-We only consider papers that are written in English, published in conferences or journals, and within the time frame of 2019 and 2024.
-The published papers should describe using the Rust programming language for either developing a new kernel, extending an existing kernel, or authoring drivers.
-We will include any type of kernel architecture including monolithic kernel, microkernel, or unikernel in both the embedded and non-embedded space as long as the paper is using the Rust programming language in some way.
-Papers that describe solutions that reside 100\% in user space will be excluded.
+Following the guidelines outlined by Kitchenham and Charters~\cite{Stuart2007-cc} we set the
+inclusion and exclusion criteria based on our research questions outlined in
+section~\ref{sec:researchQuestions}. We only consider papers that are written in English, published
+in conferences or journals, and within the time frame of 2019 and 2024. The published papers should
+describe using the Rust programming language for either developing a new kernel, extending an
+existing kernel, or authoring drivers. We will include any type of kernel architecture including
+monolithic kernel, microkernel, or unikernel in both the embedded and non-embedded space as long as
+the paper is using the Rust programming language in some way. Papers that describe solutions that
+reside 100\% in user space will be excluded. Early efforts to use Rust in kernel development
+were impacted by changes and updates in both the Rust language and compiler and relied on unstable
+features and thus are excluded from this review.
 
@@
 
-Once we had our initial collection of papers from the databases we merged all duplicate papers into one record and then started our first pass which consisted of reading the title and abstract and either marking the paper as \textit{include} or \textit{exclude}.
-Once  the initial screening was completed we were left with 26 potential papers that we needed an in depth reading.
+Once we had our initial collection of papers from the databases we merged all duplicate papers into
+one record and then started our first pass which consisted of reading the title and abstract and
+either marking the paper as \textit{include} or \textit{exclude}. Once the initial screening was
+completed we were left with 26 potential papers that we needed an in depth reading with an
+additional 2 papers found during forward and backward snowballing.
 
@@
 
-Table~\ref{tab:RQ1} summarizes our findings regarding what existing approaches researchers are using to integrate rust into the kernel.
+The integration of Rust into operating system kernel development has garnered significant attention
+due to Rust's promise of memory safety and concurrency guarantees. Several approaches and
+methodologies have emerged to leverage Rust for kernel development, each addressing different
+aspects of kernel functionality and integration.  We found several approaches that researches are
+taking to integrating Rust into kernel development.  The first approach is a greenfield approach
+were the operating system architecture is redesigned and built from the ground up in order to fully
+leverage the Rust programming language. The second approach is the incremental integration of Rust
+into existing kernels, such as the Linux
+kernel~\cite{The_kernel_development_community_undated-iw}. This methodology involves gradually
+rewriting or augmenting specific components of the kernel with Rust while maintaining compatibility
+with the existing C codebase. Table~\ref{tab:RQ1} summarizes our findings regarding what existing
+approaches researchers are using to integrate rust into the kernel.
 
-Boos et al.~\cite{Boos2020-zh} is about an experimental operating system that is aimed at addressing the state spill problem.
-State spill happens when a single service in an operating system can harbor a state change, induced by interacting with other services, that can eventually cause a system crash or system instability.
-The operating system is built in Rust and leverages the properties of Rust to construct a safer system.
-Theseus operates in a single address space and single privilege level and uses properties of the Rust programming language to realize isolation instead of relying on hardware.
-The design of the OS uses a novel cell based structure where ownership of memory and resources is enforced by the compiler.
-In short the Theseus OS is an operating system that was designed to match the Rust language instead of the more traditional route of matching the hardware.
+For the first approach, Boos et al.~\cite{Boos2020-zh} created an experimental operating system
+named Theseus that operates in a single address space and single privilege level and uses properties
+of the Rust programming language to realize isolation instead of relying on hardware. This novel
+approach pushes some of the roles that a traditional operating system would take onto the
+compiler. The design of Theseus uses a novel cell based structure where ownership of memory and
+resources are enforced by the compiler, thus avoiding these checks during runtime to get improved
+performance. In addition to memory safety, Theseus has a goal of addressing the state spill
+problem. State spill happens when a single service in an operating system can harbor a state change,
+induced by interacting with other services. That state change can eventually cause a system crash or
+system instability at some point in the future making it very difficult to track down the problem
+due to the separation between when the problem occurred and when it was observed. Theseus OS is an
+operating system that was designed to match the Rust language instead of the more traditional route
+of matching the hardware.
 
-Chen X et al.~\cite{Chen2023-wb} primary focused on developing a micro-kernel in Rust that could be formally verified by using both the liner type system or rust and a SMT solver.
-The new micro-kernel named Atmosphere follows the tradition of other micro-kernels by pushing most kernel functionally to users-space thus limiting the surface area that needs to be proved.
-The authors were able to get a 7.5:1 proof-to-code ratio which is higher that other formerly verified microkernels SeL4 and CeriKOS, which both have proof to code ratio of 19:1 and 20:1 respectively.
+While single address space operating systems present a unique opportunity to push more
+responsibility into the compiler a more traditional micro-kernel implementation was taken by Chen X
+et al.~\cite{Chen2023-wb} and Liang et al.~\cite{Liang2021-bo}. By pushing most kernel functionally
+to user space the authors limited the surface area that needs to be proved. Chen X et al. took
+their implementation one step further by formally verifying their new micro-kernel named Atmosphere
+by using both the liner type system of Rust in combination with a SMT solver. Chen X et al. were
+able to get a 7.5:1 proof-to-code ratio which is higher that other formerly verified micro-kernels
+SeL4 and CeriKOS, which both have proof to code ratio of 19:1 and 20:1 respectively. While this
+approach was novel drivers running in userspace are still not formally verified and don't have to be
+written in Rust.
 
-Lankes et al.~\cite{Lankes2019-cm} detailed their experiences with developing a Unikernel in Rust.
-Originally written in C and called HermitCore the authors ported the Kernel to the Rust programming language in order to take advantage of Rust's memory safe features.
-The new Unikernel consists of only 3.27\% unsafe Rust with the rest of the code base consisting of safe Rust.
+Unikernels present an ideal surface for prototyping Rust. Unikernels are designed to do one thing
+and one thing only, thus the limiting the complexity that a general purpose operating system has to
+deal with. This reduced complexity was explored by Lankes et al.~\cite{Lankes2019-cm}. The authors
+detailed their experiences with rewriting an existing Unikernel, HermitCore, in Rust. The new
+Unikernel named RustyHermit consists of only 3.27\% unsafe Rust with the rest
+of the code base consisting of safe Rust dramatically decreasing the surface area where memory errors
+could originate from.
 
-Li and Sato~\cite{Li2024-yb} explored using Rust to write an exokernel named W-Kernel.
-The authors proposed a novel architecture for an OS by embedding a WebAssembly(WASM) runtime into the kernel what can run programs written in any language that can compile down to WASM.
+Finally, Li and Sato~\cite{Li2024-yb} explored using Rust to write an exokernel named W-Kernel. The authors
+proposed a novel architecture for an OS by embedding a WebAssembly(WASM) runtime into the kernel
+what can run programs written in any language that can compile down to WASM.
 
-Liang et al.~\cite{Liang2021-bo} developed a new operating system named Rustpi with a micro-kernel architecture implemented in Rust to explore how modern language features can help with building reliable systems.
-
-\begin{table}
+\begin{table*}
     \begin{tabular}{||l|l|l||}
@@
     \label{tab:RQ1}
-\end{table}
+\end{table*}
 
 \subsection{RQ2  What are the performance implications of using Rust for operating system
-      kernel development in terms of throughput, latency, and resource utilization?}
+      kernel development in terms of throughput, and  latency?}
 
-Comparing performance, throughput, latency, and resource utilization between different systems is very difficult.
-For example several of the operating systems written in Rust such as Tock are a complete rethinking of how an operating system is designed and Rust for Linux (RFL) is still relatively new and has not had the benefit of decades of incremental performance improvements like components that are written in C.
-Additionally, direct comparisons between a Monolithic kernel, Micro-kernel, and Unikernel, or comparisons between a Real Time Operation System (RTOS) and a General Purpose operating System are not directly meaningful due to the vastly different goals of each system.
-Therefore, we will focus more on the overall efforts that are specific to the Rust programming language and the challenges presented.
-We summarize our findings in table~\ref{tab:RQ2}.
+Comparing performance, throughput and latency between different systems is very difficult. For
+example several of the operating systems written in Rust such as Tock are a complete rethinking of
+how an operating system is designed and Rust for Linux (RFL) is still relatively new and has not had
+the benefit of decades of incremental performance improvements like components that are written in
+C. Additionally, direct comparisons between a Monolithic kernel, Micro-kernel, and Unikernel, or
+comparisons between a Real Time Operation System (RTOS) and a General Purpose operating System are
+not directly meaningful due to the vastly different goals of each system. Therefore, we will focus
+more on the overall efforts that are specific to the Rust programming language and the challenges
+presented. We summarize our findings in table~\ref{tab:RQ2}.
 
-Culic et al.~\cite{Culic2022-bk} looked at latency issues in the Tock OS.
-Tock is a new operating system that is designed to run on embedded systems but does not provide Real time capabilities.
-The authors attempted to add real time capabilities by integrating eBPF into the Tock kernel to improve the interrupt handlers response time.
-The Authors found that early work (still in the prototype stage) lowers the response times of the system and makes interrupt response times 3x.
+Culic et al.~\cite{Culic2022-bk} looked at latency issues in the Tock OS. Tock is a new operating
+system that is designed to run on embedded systems but does not provide Real time capabilities. The
+authors attempted to add real time capabilities by integrating eBPF into the Tock kernel to improve
+the interrupt handlers response time. The Authors found that early work (still in the prototype
+stage) lowers the response times of the system and makes interrupt response times 3x.
 
-Gonzalez et al.~\cite{Gonzalez2023-ek} explored using the Rust for Linux Project to implement a native UDP driver in Rust in order to explore the performance the Rust programming language.
-The authors were able to get a basic driver working with performance only slightly slower than C using the Rust for Linux (RFL) project.
-The RFL project is still to immature to get a full driver up and running but is at a stage were we researchers can start experimenting with different approaches.
+Gonzalez et al.~\cite{Gonzalez2023-ek} explored using the Rust for Linux Project to implement a
+native UDP driver in Rust in order to explore the performance the Rust programming language. The
+authors were able to get a basic driver working with performance only slightly slower than C using
+the Rust for Linux (RFL) project. The RFL project is still to immature to get a full driver up and
+running but is at a stage were we researchers can start experimenting with different approaches.
 
-Li et al.~\cite{Li2024-be} explored the feasibility of using Rust in kernel space.
-The authors took an existing component, the Out of Memory (OOM) and implemented a replacement using the Rust programming language.
-The non-encapsulated interface Rust component which was almost identical to the original C component only introduced a 0.7\% overhead.
-The encapsulated Rust component on the other hand added a 3\% performance overhead.
-Another important aspect of the kernel is size, we must be careful to keep the size of the Rust component as close as we can to the C component so we can still run on all the same hardware.
-The authors found that there was only a 0.06\% size increase when compared to the original C implementation.
+Li et al.~\cite{Li2024-be} explored the feasibility of using Rust in kernel space. The authors took
+an existing component, the Out of Memory (OOM) and implemented a replacement using the Rust
+programming language. The non-encapsulated interface Rust component which was almost identical to
+the original C component only introduced a 0.7\% overhead. The encapsulated Rust component on the
+other hand added a 3\% performance overhead. Another important aspect of the kernel is size, we must
+be careful to keep the size of the Rust component as close as we can to the C component so we can
+still run on all the same hardware. The authors found that there was only a 0.06\% size increase
+when compared to the original C implementation.
 
@@
     3 & Latency & \cite{Culic2022-bk} \\
-    4 & Resource Utilization & 0 \\
     \hline
@@
         \hline
-        5 & C Interop \cite{Miller2019-xm} &
+        5 & C Interop \cite{Miller2019-xm}, \cite{Li2021-xo} &
         \begin{itemize}
@@
 
-Ayers et al.~\cite{Ayers2022-sf} focused on the size of rust binaries in the embedded environment.
-Their work primary focused on the size of binaries produced using the embedded OS Tock.
-In addition to the identified causes of binary growth the authors have the following 5 recommendations when using Rust in a size constrained environment:
+The development of operating system kernels in Rust introduces several unique challenges and
+limitations compared to more traditional languages like C. These challenges arise from Rust’s strict
+safety guarantees, its relatively recent adoption in systems programming, and the inherent
+complexities of kernel development. This section discusses these challenges, limitations, and the
+lessons learned from various projects and research efforts.
+
+A primary concern for programming languages that are used for operating system development are the
+size of the binary especially in an embedded environment.  Ayers et al.~\cite{Ayers2022-sf} focused
+on reducing size of binaries produced by the rust compiler while working with Tock OS. They were
+able to identify several  causes of binary growth that are specific to the Rust programming language
+and have the following 5 recommendations when using Rust in a size constrained environment:
 \begin{itemize}
@@
 
-Burtsev et al.~\cite{Burtsev2021-mh} explores what is missing in the Rust programming language to help solve the isolation problem in an operating system.
-Currently Rust lacks the ability to express isolation in the heap without external support.
-The RedLeaf experiential operating system relied on an complex interface definition language (IDL) to enforce isolated heaps.
-The paper enumerates several properties of the Rust Language that could help with isolation with regards to operating system development that are detailed in table~\ref{tab:RQ3}.
-The authors argue that with these changes or inclusions to the Rust programming language developing operating system kernels would be much easier and safer.
+Burtsev et al.~\cite{Burtsev2021-mh} explored what is missing in the Rust programming language to
+help solve the isolation problem in an operating system. Currently Rust lacks the ability to express
+isolation in the heap without external support. For example, the RedLeaf experiential operating
+system which is written in Rust relied on an complex interface definition language (IDL) to enforce
+isolated heaps. The paper enumerates several properties of the Rust Language that could help with
+isolation with regards to operating system development that are detailed in table~\ref{tab:RQ3}. The
+authors argue that with these changes or inclusions to the Rust programming language developing
+operating system kernels would be much easier and safer.
 
-Klimt et al.~\cite{Klimt2023-ob} details the lessons learned and challenges when implementing Theseus, an operating system written in Rust.
-First they found that it is impossible to write a complete operating system in 100\% safe rust.
-For example when writing a memory management system raw pointers must be used in order to modify the hardware.
-Despite the limitations imposed by the hardware the authors describe how Theseus leveraged intralingual design to maximize the compilers role in enforcing correctness.
-By leveraging Rusts type system and borrow checker memory safety and correct ownership transfer can be achieved at a higher level than what could be done in C.
-The authors also detailed some of the limitations of intralingual design such as not being as expressive as many other formal verification techniques due to the limited invariant that can be enforced by the type system.
-One of the most important lessons learned was the insight that a linear type system itself cannot guarantee uniqueness of the resource represented, such when a memory resources may overlap.
-The authors introduce the idea of using a hybrid approach of verification where they leverage both the linear type system and an SMT solver.
-The authors also explore the bootstrapping problem with rust systems.
-How do we ensure that ownership of memory is sound in the operating system itself?
-In operating systems written in C the kernel provides the ownership root to applications running on top, finding a new ownership root is a open research question with regards to using Rust for operating system development.
+Klimt et al.~\cite{Klimt2023-ob} details the lessons learned and challenges when implementing
+Theseus. First they found that it is impossible to write a complete operating system in 100\% safe
+Rust. For example when writing a memory management system raw pointers must be used in order to
+modify the hardware. Despite the limitations imposed by the hardware the authors describe how
+Theseus leveraged intralingual design to maximize the compilers role in enforcing correctness. By
+leveraging Rusts type system and borrow checker memory safety and correct ownership transfer can be
+achieved at a higher level than what could be done in C. The authors also detailed some of the
+limitations of intralingual design such as not being as expressive as many other formal verification
+techniques due to the limited invariant that can be enforced by the type system. One of the most
+important lessons learned was the insight that a linear type system itself cannot guarantee
+uniqueness of the resource represented, such when a memory resources may overlap. The authors
+introduce the idea of using a hybrid approach of verification where they leverage both the linear
+type system and an SMT solver. The authors also explore the bootstrapping problem with rust
+systems. How do we ensure that ownership of memory is sound in the operating system itself? In
+operating systems written in C the kernel provides the ownership root to applications running on
+top, finding a new ownership root is a open research question with regards to using Rust for
+operating system development.
+
+\section{Threats to Validity}
+
+In conducting this systematic literature review on the use of Rust for operating system development,
+several potential threats to validity must be acknowledged. These threats are categorized into
+internal, external, and conclusion validity, each presenting distinct challenges to the robustness
+and reliability of our findings.
+
+Definition Ambiguity: The term "Kernel development in Rust" encompasses a wide range of activities,
+from minor module development to complete subsystem rewrites. Variations in how researchers and
+developers interpret this integration could lead to inconsistent findings. Selection Criteria: The
+inclusion and exclusion criteria for selecting relevant studies might inadvertently bias the review
+towards certain types of research, such as those reporting successful integration while
+underrepresenting studies detailing challenges and failures. Compared to C Rust is a brand new
+language that is still finding its way in the world.
+
+Internal Validity: Internal validity refers to the degree to which the review accurately establishes
+causal relationships. Potential threats include:
+
+\begin{itemize}
+  \item Publication Bias: Studies demonstrating positive outcomes of using Rust for kernel
+    development may be more likely to be published and cited, leading to an overestimation of its
+    benefits.
+  \item Quality of Studies: The methodological rigor of the included studies varies, and some may
+    suffer from design flaws or biases that were not adequately controlled for, thus affecting the
+    reliability of their findings.
+\end{itemize}
+
+External Validity: External validity pertains to the generalizability of the review's findings
+beyond the specific context of the reviewed studies. Threats to external validity include:
+
+\begin{itemize}
+  \item Context-Specific Findings: Many studies focus on specific kernel modules or use-cases, which
+    may not be representative of the broader kernel environment. The success of Rust in isolated
+    components does not necessarily translate to the entire kernel.
+  \item Temporal Changes: The rapidly evolving nature of both the Rust programming language and the
+    Linux kernel means that findings from older studies may no longer be applicable, as improvements
+    and changes in both domains can alter the landscape significantly.
+\end{itemize}
+
+Conclusion Validity: Conclusion validity involves the degree to which the conclusions drawn from the
+review are credible and supported by the data. Potential threats include:
+
+\begin{itemize}
+  \item Heterogeneity of Studies: The diverse methodologies, metrics, and contexts of the included studies
+    can lead to challenges in synthesizing findings and drawing unified conclusions. For example, it
+    is very difficult to directly compare a micro-kernel, Unikernel, and hybrid Kernel.
+  \item Reviewer Bias: Personal biases of the reviewers in interpreting data and making judgments
+    about study quality and relevance may skew the results
+\end{itemize}
+
+Efforts were made to minimize this through predefined criteria and multiple reviewers, but some
+degree of subjectivity is inevitable.  In addressing these threats, we employed rigorous methods for
+study selection, data extraction, and analysis, and we remain transparent about the
+limitations. Future reviews could mitigate these threats further by incorporating a broader range of
+sources, using more precise definitions, and employing meta-analytic techniques to quantitatively
+synthesize findings. Despite these threats, our review provides a valuable synthesis of the current
+state of research on the use of Rust in kernel development, highlighting both its potential benefits
+and the challenges that need to be addressed.
 
@@
 
-Summary of key findings and insights obtained from the systematic literature review.
-Concluding remarks on the significance of memory-safe programming languages in mitigating security vulnerabilities.
-Closing thoughts on the importance of minimizing unsafe blocks during code migration.
+The integration of Rust into  kernel space represents a significant evolution in operating system
+development, promising to address long-standing issues related to memory safety and system
+reliability. This systematic literature review has synthesized the current state of research and
+practice concerning the use of Rust in multiple types of kernels including the venerable Linux
+kernel, highlighting both the progress made and the challenges that remain.
 
-\section{Threats to Validity}
+Our review indicates that Rust's strong guarantees of memory safety, enabled by its ownership model
+and strict compile-time checks, offer a compelling advantage over traditional C-based kernel
+development. These features have the potential to reduce common vulnerabilities such as buffer
+overflows and use-after-free errors, which are prevalent in C and have historically led to critical
+security exploits. Several case studies and prototype implementations have demonstrated that Rust
+can be successfully integrated into the kernel, providing safer interfaces and reducing the
+incidence of memory-related bugs without incurring significant performance penalties.
+
+However, the adoption of Rust for kernel development is not without its challenges. Truly the
+biggest challenge is existing codebases and mountains of legacy code. With extensive use of C and
+reliance on specific C idioms and low-level programming techniques researchers have substantial
+integration hurdles to overcome. Any efforts to rewrite substantial portions of an existing kernel
+such as Linux in Rust are constrained by the need for interoperability with existing C code and the
+necessity to maintain the kernel's performance characteristics. Furthermore, the operating system
+development community must reach consensus on Rust's role and ensure that sufficient tooling,
+documentation, and support are available for developers.
+
+In conclusion, while Rust's incorporation into the kernel space is still in its nascent stages, the
+initial results are promising. The potential for enhanced security and stability aligns with the
+long-term goals of kernel development, and continued research, coupled with practical implementation
+efforts, will be crucial in realizing these benefits. Future work should focus on addressing
+integration challenges, refining interoperability mechanisms, and expanding the body of empirical
+evidence on Rust's impact within the kernel environment. As the community navigates these
+challenges, the evolution of Rust in the kernel may well mark a transformative period in the pursuit
+of safer, more reliable operating systems.
+
 \balance

### 2024-05-30 paper.tex
Updated rq3
Update on Overleaf.

@@
-\documentclass[sigconf]{acmart}
+%\documentclass[sigconf]{acmart}
+\documentclass[sigconf,review,anonymous]{acmart}
 \usepackage{listings, listings-rust}
@@
 \acmDOI{XXXXXXX.XXXXXXX}
-\acmSubmissionID{123-A56-BU3}
+%\acmSubmissionID{123-A56-BU3}
 
 \begin{document}
-\title{Kernel Development Using Rust: A Systematic Literature Review}
+%\title{Kernel Development Using Rust: A Systematic Literature Review}
+\title{Rusty Linux: Advances in Rust for Linux Kernel Development}
 
-\author{Anonymous Authors}
+%\author{Anonymous Authors}
 
@@
   at enhancing system security and reliability by leveraging Rust's strong memory safety
-  guarantees. This systematic literature review examines the current state of research and practice
-  regarding the use of Rust in the kernel development.
-  \textit{Objective:} We aim to find the current edge of research regarding using Rust in Kernel
-  development so we can push the research forward and reduce the number of memory safety
+  guarantees. 
+  %This systematic literature review examines the current state of research and practice regarding the use of Rust in the kernel development.
+  \textit{Objective:} We aim to find the current advances in using Rust in Kernel
+  development to reduce the number of memory safety
   vulnerabilities in one of the most critical pieces of software that underpins all modern
@@
   \textit{Method:} By analyzing a broad spectrum of studies, we identify the advantages Rust offers,
-  such as reducing vulnerabilities like buffer overflows and use-after-free errors, which are
-  prevalent in C-based kernel development. The review also highlights the challenges faced,
-  including integration hurdles with existing C codebases, performance considerations, and the need
-  for community consensus on Rust's adoption.
+  %such as reducing vulnerabilities like buffer overflows and use-after-free errors, which are
+  %prevalent in C-based kernel development. The review also 
+  highlight the challenges faced,
+  %including integration hurdles with existing C codebases, performance considerations, 
+  and emphasise the need for community consensus on Rust's adoption.
   \textit{Results:} Our findings suggest that while the initial implementations of Rust in the kernel
-  shows promising results in terms of safety and stability, significant challenges remain. These
+  show promising results in terms of safety and stability, significant challenges remain. These challenges
   include achieving seamless interoperability with existing kernel components, maintaining
   performance, and ensuring adequate support and tooling for developers.
-  \textit{Conclusions:} In conclusion, this review provides a comprehensive overview of the
-  potential and challenges of incorporating Rust into kernel development. It underscores the need for
+  \textit{Conclusions:} This study 
+  %provides a comprehensive overview of the
+  %potential and challenges of incorporating Rust into kernel development. It 
+  underscores the need for
   continued research and practical implementation efforts to fully realize the benefits of Rust. By
@@
 The 1995 movie Hackers~\cite{Wikipedia_contributors2024-zr} prescient predictions regarding the ease
-of breaking into computing systems in cyberspace have came to fruition. The White House Office of
+of breaking into computing systems in cyberspace have come to fruition. The White House Office of
 the National Cyber Director (ONCD) released a report calling for the technical community to
-proactively reduce the attack surface in cyberspace with a two pronged approach. First we need to
+proactively reduce the attack surface in cyberspace with a two-pronged approach. First, we need to
 address the root cause of many of the most heinous cyber attacks, memory unsafe programming
 languages~\cite{United_States_Gov2024-pp}. Second, we need to establish better cybersecurity quality
-metrics so we can have a better understanding of the cyber security landscape.
+metrics so we can have a better understanding of the cybersecurity landscape.
 
 In the ever-evolving landscape of software development, the reliability and security of computer
-systems stands as a paramount concern for all parties involved. Modern software is constructed by
+systems stand as a paramount concern for all parties involved. Modern software is constructed by
 building ever more complex abstractions, one on top of the other. Thus, if we aim to have a secure
-system we must start to peel back all the layers and tackle one of the fundamental abstractions in
-computer science, the programming language. Programming languages that provide and enforce memory
+system, we must start to peel back all the layers and tackle one of the fundamental abstractions in
+computer science: the programming language. Programming languages that provide and enforce memory
 safety eliminate whole classes of bugs such as buffer overflows, dangling pointers, and memory leaks
@@
 landscape of operating system development. This shift prompts a comprehensive evaluation of both the
-opportunities and challenges associated with using Rust in such a foundational component of modern
+opportunities and challenges associated with using Rust as such a foundational component of modern
 computing. This systematic literature review aims to synthesize existing research and practical
@@
 for integrating Rust into one of the most fundamental areas that are typically dominated by unsafe
-languages the operating system kernel. Our study aims to provide a comprehensive overview of
+languages, the operating system kernel. Our study aims to provide a comprehensive overview of
 existing research, identify gaps, and suggest future directions in this domain. Through a rigorous
@@
 
-For our research methodolgy we followed the advice of Kitchenham and Charters~\cite{Stuart2007-cc}
+For our research methodology we followed the advice of Kitchenham and Charters~\cite{Stuart2007-cc}
 and divided our review into the 3 discrete phases, planning the review, conducting the review, and
@@
 complexities of kernel development. This section discusses these challenges, limitations, and the
-lessons learned from various projects and research efforts.
+lessons learned from various projects and research efforts.  We summarize our findings in
+table~\ref{tab:RQ3}.
 

### 2024-05-30 paper.tex
Update on Overleaf.

@@
 
-  \textit{Context:} The integration of Rust into kernel development is a transformative endeavor aimed
+  \textit{\textbf{Context:}} The integration of Rust into kernel development is a transformative endeavor aimed
   at enhancing system security and reliability by leveraging Rust's strong memory safety
@@
   %This systematic literature review examines the current state of research and practice regarding the use of Rust in the kernel development.
-  \textit{Objective:} We aim to find the current advances in using Rust in Kernel
+  \textit{\textbf{Objective:}} We aim to find the current advances in using Rust in Kernel
   development to reduce the number of memory safety
@@
   applications.
-  \textit{Method:} By analyzing a broad spectrum of studies, we identify the advantages Rust offers,
+  \textit{\textbf{Method:}} By analyzing a broad spectrum of studies, we identify the advantages Rust offers,
   %such as reducing vulnerabilities like buffer overflows and use-after-free errors, which are
@@
   and emphasise the need for community consensus on Rust's adoption.
-  \textit{Results:} Our findings suggest that while the initial implementations of Rust in the kernel
+  \textit{\textbf{Results:}} Our findings suggest that while the initial implementations of Rust in the kernel
   show promising results in terms of safety and stability, significant challenges remain. These challenges
@@
   performance, and ensuring adequate support and tooling for developers.
-  \textit{Conclusions:} This study 
+  \textit{\textbf{Conclusions:}} This study 
   %provides a comprehensive overview of the
@@
 computer science: the programming language. Programming languages that provide and enforce memory
-safety eliminate whole classes of bugs such as buffer overflows, dangling pointers, and memory leaks
+safety eliminate whole classes of bugs, such as buffer overflows, dangling pointers, and memory leaks
 which have been implicated in a myriad of security vulnerabilities and system crashes.

### 2024-06-01 paper.tex
Update on Overleaf.

@@
 of a garbage collector, making it an ideal candidate for kernel development where safety and
-performance are critical.
+performance are critical.~\cite{}
 

### 2024-06-02 paper.tex
Update on Overleaf.

--- 2024-05-30 paper.tex
@@
 of a garbage collector, making it an ideal candidate for kernel development where safety and
-performance are critical.
+performance are critical~\cite{noauthor_undated-id}.
 

### 2024-06-02 paper.tex
Update on Overleaf.

--- 2024-05-30 paper.tex
@@
 vulnerabilities, including buffer overflows and use-after-free errors, which have plagued operation
-system development for decades~\cite{noauthor_undated-bf}. Addressing these vulnerabilities is
+system development for decades~\cite{noauthor_2019-bf}. Addressing these vulnerabilities is
 paramount to enhancing the security and reliability of all operating systems.
@@
 of a garbage collector, making it an ideal candidate for kernel development where safety and
-performance are critical.
+performance are critical~\cite{noauthor_2024-id}.
 

### 2024-06-02 paper.tex
Update on Overleaf.

@@
 of a garbage collector, making it an ideal candidate for kernel development where safety and
-performance are critical~\cite{noauthor_2024-id}.
+performance are critical~\cite{}.
 

### 2024-06-03 paper.tex
Update on Overleaf.

@@
 the National Cyber Director (ONCD) released a report calling for the technical community to
-proactively reduce the attack surface in cyberspace with a two-pronged approach. First, we need to
-address the root cause of many of the most heinous cyber attacks, memory unsafe programming
-languages~\cite{United_States_Gov2024-pp}. Second, we need to establish better cybersecurity quality
+proactively reduce the attack surface in cyberspace with a two-pronged approach~\cite{United_States_Gov2024-pp}. First, we need to
+address the root cause of many of the most heinous cyber attacks: memory-unsafe programming
+languages. Second, we need to establish better cybersecurity quality
 metrics so we can have a better understanding of the cybersecurity landscape.
@@
 vulnerabilities, including buffer overflows and use-after-free errors, which have plagued operation
-system development for decades~\cite{noauthor_2019-bf}. Addressing these vulnerabilities is
+system development for decades~\cite{MSRC_Team2019-bf}. Addressing these vulnerabilities is
 paramount to enhancing the security and reliability of all operating systems.
@@
 of a garbage collector, making it an ideal candidate for kernel development where safety and
-performance are critical~\cite{}.
+performance are critical~\cite{Klabnik2024-id}.
 
@@
 opportunities and challenges associated with using Rust as such a foundational component of modern
-computing. This systematic literature review aims to synthesize existing research and practical
+computing. Therefore, we aim to synthesize existing research and practical
 experiences related to the use of Rust in the kernel, providing a detailed understanding of
@@
 
-This paper presents a systematic literature review (SLR) focusing on strategies and methodologies
+This reflection paper presents a systematic literature review (SLR) focusing on strategies and methodologies
 for integrating Rust into one of the most fundamental areas that are typically dominated by unsafe
-languages, the operating system kernel. Our study aims to provide a comprehensive overview of
+languages, the operating system kernel. We aim to provide a comprehensive overview of
 existing research, identify gaps, and suggest future directions in this domain. Through a rigorous
@@
 
-For our research methodology we followed the advice of Kitchenham and Charters~\cite{Stuart2007-cc}
-and divided our review into the 3 discrete phases, planning the review, conducting the review, and
+For our research methodology, we followed the Kitchenham and Charters~\cite{Stuart2007-cc} methodology to conduct SLR in software engineering.
+We divided our review into three discrete phases, planning the review, conducting the review, and
 reporting the review results. The following sections detail our review process.
@@
 
-Before we start researching we must first confirm the need for a SLR. The recent report released by
+To begin our study, we must first confirm the need for a SLR. The recent report released by
 ONCD~\cite{United_States_Gov2024-pp} has conveniently done this job for us by compiling a report
-detailing the need for research in the domain of memory safety. While the ONCD report detailed a two
-pronged approach, for this SLR we will be focusing on a memory safe programming language,
-specifically Rust. While there are many modern memory safe programming languages available for
-software developers to use Rust is one of the few languages that is feasible to use when developing
-operating system kernels due to is lack of runtime and garbage collector. Even the Linux kernel is
+detailing the need for research in the domain of memory safety.
+While the ONCD report detailed a two-pronged approach, for this reflection paper, we will be focusing on a memory-safe programming language,
+specifically Rust. While there are many modern memory-safe programming languages available for
+software developers to use, Rust is one of the few languages that is feasible to use when developing
+operating system kernels due to its lack of runtime and garbage collector. Even the Linux kernel is
 in the early stages of adding real support for
@@
 
-\subsubsection{Research Questions}
+\subsubsection{\textbf{Research Questions}}
 \label{sec:researchQuestions}
@@
 \begin{itemize}
-    \item \textbf{RQ1:} What are the existing approaches and methodologies for implementing
+    \item \textbf{RQ1:} What are the existing approaches for implementing
       operating system kernels in Rust?
     \item \textbf{RQ2:} What are the performance implications of using Rust for operating system
-      kernel development in terms of throughput, latency, and resource utilization?
+      kernel development?
+      %in terms of throughput, latency, and resource utilization?
     \item \textbf{RQ3:} What are the major challenges, limitations, and lessons learned when
-      developing operating system kernels in Rust, particularly in comparison to other languages?
+      developing operating system kernels in Rust?% particularly in comparison to other languages?
 \end{itemize}
 
-\subsubsection{Search Strategy:}
+\subsection{Data Collection}
 
-We employed a robust multi-step search strategy across three digital databases, ACM Digital Library,
-IEEE Xplore, and Scopus in order to find all the current research that has been done. These
-databases cover a wide range of current software engineering and computer science literature. We
-leveraged the advanced search features of all three databases to search the titles and abstracts for
+\subsubsection{\textbf{Search Strategy.}}
+
+We employed a robust multi-step search strategy across three databases: ACM Digital Library,
+IEEE Xplore, and Scopus to find all the current research. 
+%These databases cover a wide range of current software engineering and computer science literature. 
+We leveraged the advanced search features of all three databases to search the titles and abstracts for
 keywords using boolean search operators.
 
-\subsubsection{Search criteria:}
+\subsubsection{\textbf{Search criteria.}}
 
-We search all three databases with the keywords listed in table~\ref{tab:keywords} between January
-1, 2019 and April 1, 2024. While both ACM and IEEE had mutually exclusive results Scopus had 44
+We searched all three databases with the keywords listed in table~\ref{tab:keywords} between January
+1, 2019 and April 1, 2024. While both ACM and IEEE had mutually exclusive results, Scopus had 44
 duplicates that needed to be removed. We then used Google Scholar for forward and backward
 snowballing. We choose to look at only the previous 5 years of research in order to capture the
-bleeding edge research that is currently being done.
+bleeding-edge research that is currently being done.
 
@@
 
-\subsubsection{Inclusion and exclusion criteria:}
+\subsubsection{\textbf{Inclusion and exclusion criteria.}}
 
@@
 describe using the Rust programming language for either developing a new kernel, extending an
-existing kernel, or authoring drivers. We will include any type of kernel architecture including
-monolithic kernel, microkernel, or unikernel in both the embedded and non-embedded space as long as
+existing kernel, or authoring drivers. We include any type of kernel architecture, including
+monolithic kernel, microkernel, or unikernel in both the embedded and non-embedded space, as long as
 the paper is using the Rust programming language in some way. Papers that describe solutions that
@@
 were impacted by changes and updates in both the Rust language and compiler and relied on unstable
-features and thus are excluded from this review.
+features, which are thus excluded from this review.
 
-\subsection{Conducting the review}
+\subsection{Data Analysis}
 
-Once we had our initial collection of papers from the databases we merged all duplicate papers into
-one record and then started our first pass which consisted of reading the title and abstract and
+Once we had our initial collection of papers from the databases, we merged all duplicate papers into
+one record and then started our first pass, which consisted of reading the title and abstract and
 either marking the paper as \textit{include} or \textit{exclude}. Once the initial screening was
-completed we were left with 26 potential papers that we needed an in depth reading with an
-additional 2 papers found during forward and backward snowballing.
+completed we were left with 26 potential papers that we needed an in-depth reading with an
+additional 2 papers were found during forward and backward snowballing.
 

### 2024-06-05 paper.tex
Update on Overleaf.

@@
 compile-time checks effectively prevent common programming errors that lead to security
-vulnerabilities. What makes Rust unique is the fact that it accomplishes these tasks without the use
-of a garbage collector, making it an ideal candidate for kernel development where safety and
-performance are critical~\cite{Klabnik2024-id}.
+vulnerabilities~\cite{Klabnik2024-id}.
+What makes Rust unique is the fact that it accomplishes these tasks without the use of a garbage collector, making it an ideal candidate for kernel development where safety and performance are critical.
 
@@
 \begin{itemize}
-    \item \textbf{RQ1:} What are the existing approaches for implementing
-      operating system kernels in Rust?
-    \item \textbf{RQ2:} What are the performance implications of using Rust for operating system
-      kernel development?
+    \item \textbf{RQ1:} What are the existing approaches for implementing operating system kernels in Rust?
+    \item \textbf{RQ2:} What are the performance implications of using Rust for operating system kernel development?
       %in terms of throughput, latency, and resource utilization?
-    \item \textbf{RQ3:} What are the major challenges, limitations, and lessons learned when
+    \item \textbf{RQ3:} What are the major challenges and limitations when
       developing operating system kernels in Rust?% particularly in comparison to other languages?
+    \item \textbf{RQ4:} What are the lessons learned when developing operating systems kernels in Rust?
 \end{itemize}
@@
 section~\ref{sec:researchQuestions}. We only consider papers that are written in English, published
-in conferences or journals, and within the time frame of 2019 and 2024. The published papers should
+in conferences, journals, workshops, and within the time frame of 2019 and 2024. The published papers should
 describe using the Rust programming language for either developing a new kernel, extending an
@@
 
-In this section we present our findings. We address each research question from \textbf{RQ1} - \textbf{RQ3}.
+In this section, we present our findings. We address each research question from \textbf{RQ1} - \textbf{RQ3}.
 
-\subsection{RQ1 What are the existing approaches and methodologies for implementing operating system kernels in Rust?}
+\subsection{RQ1: Existing Approaches}
 
@@
 methodologies have emerged to leverage Rust for kernel development, each addressing different
-aspects of kernel functionality and integration.  We found several approaches that researches are
-taking to integrating Rust into kernel development.  The first approach is a greenfield approach
-were the operating system architecture is redesigned and built from the ground up in order to fully
+aspects of kernel functionality and integration.  We found several approaches that researchers are
+taking to integrate Rust into kernel development. The first approach is a greenfield approach
+where the operating system architecture is redesigned and built from the ground up in order to fully
 leverage the Rust programming language. The second approach is the incremental integration of Rust
@@
 approach pushes some of the roles that a traditional operating system would take onto the
-compiler. The design of Theseus uses a novel cell based structure where ownership of memory and
-resources are enforced by the compiler, thus avoiding these checks during runtime to get improved
+compiler. The design of Theseus uses a novel cell-based structure where ownership of memory and
+resources is enforced by the compiler, thus avoiding these checks during runtime to get improved
 performance. In addition to memory safety, Theseus has a goal of addressing the state spill
-problem. State spill happens when a single service in an operating system can harbor a state change,
+problem. State spill happens when a single service in an operating system can harbor a state change
 induced by interacting with other services. That state change can eventually cause a system crash or
@@
 responsibility into the compiler a more traditional micro-kernel implementation was taken by Chen X
-et al.~\cite{Chen2023-wb} and Liang et al.~\cite{Liang2021-bo}. By pushing most kernel functionally
-to user space the authors limited the surface area that needs to be proved. Chen X et al. took
+et al.~\cite{Chen2023-wb} and Liang et al.~\cite{Liang2021-bo}. By pushing most kernel functionality
+to user space, the authors limited the surface area that needs to be proved. Chen X et al.~\cite{Chen2023-wb} took
 their implementation one step further by formally verifying their new micro-kernel named Atmosphere
-by using both the liner type system of Rust in combination with a SMT solver. Chen X et al. were
-able to get a 7.5:1 proof-to-code ratio which is higher that other formerly verified micro-kernels
-SeL4 and CeriKOS, which both have proof to code ratio of 19:1 and 20:1 respectively. While this
+by using both the liner type system of Rust in combination with a SMT solver. They were
+able to get a 7.5:1 proof-to-code ratio, which is higher than other formerly verified micro-kernels
+SeL4 and CeriKOS both have proof code ratios of 19:1 and 20:1, respectively. While this
 approach was novel drivers running in userspace are still not formally verified and don't have to be
@@
 Unikernels present an ideal surface for prototyping Rust. Unikernels are designed to do one thing
-and one thing only, thus the limiting the complexity that a general purpose operating system has to
+and one thing only, thus limiting the complexity that a general-purpose operating system has to
 deal with. This reduced complexity was explored by Lankes et al.~\cite{Lankes2019-cm}. The authors
 detailed their experiences with rewriting an existing Unikernel, HermitCore, in Rust. The new
-Unikernel named RustyHermit consists of only 3.27\% unsafe Rust with the rest
-of the code base consisting of safe Rust dramatically decreasing the surface area where memory errors
-could originate from.
+Unikernel named RustyHermit consists of only 3.27\% unsafe Rust, with the rest
+of the code base consisting of safe Rust, dramatically decreasing the surface area where memory errors
+could originate.
 
 Finally, Li and Sato~\cite{Li2024-yb} explored using Rust to write an exokernel named W-Kernel. The authors
-proposed a novel architecture for an OS by embedding a WebAssembly(WASM) runtime into the kernel
-what can run programs written in any language that can compile down to WASM.
+proposed a novel architecture for an OS by embedding a WebAssembly(WASM) runtime into the kernel that can run programs written in any language that can compile down to WASM.
 
@@
 
-\subsection{RQ2  What are the performance implications of using Rust for operating system
-      kernel development in terms of throughput, and  latency?}
+\subsection{RQ2: Performance Implications}
 
 Comparing performance, throughput and latency between different systems is very difficult. For
-example several of the operating systems written in Rust such as Tock are a complete rethinking of
-how an operating system is designed and Rust for Linux (RFL) is still relatively new and has not had
+example, several of the operating systems written in Rust, such as Tock, are a complete rethinking of
+how an operating system is designed, and Rust for Linux (RFL) is still relatively new and has not had
 the benefit of decades of incremental performance improvements like components that are written in
 C. Additionally, direct comparisons between a Monolithic kernel, Micro-kernel, and Unikernel, or
-comparisons between a Real Time Operation System (RTOS) and a General Purpose operating System are
+comparisons between a Real Time Operation System (RTOS) and a General Purpose Operating System, are
 not directly meaningful due to the vastly different goals of each system. Therefore, we will focus
@@
 Culic et al.~\cite{Culic2022-bk} looked at latency issues in the Tock OS. Tock is a new operating
-system that is designed to run on embedded systems but does not provide Real time capabilities. The
-authors attempted to add real time capabilities by integrating eBPF into the Tock kernel to improve
-the interrupt handlers response time. The Authors found that early work (still in the prototype
-stage) lowers the response times of the system and makes interrupt response times 3x.
+system that is designed to run on embedded systems but does not provide Real-time capabilities. The
+authors attempted to add real-time capabilities by integrating eBPF into the Tock kernel to improve
+the interrupt handler's response time. The Authors found that early work (still in the prototype
+stage) lowers the response times of the system and interrupts response times 3x.
 
 Gonzalez et al.~\cite{Gonzalez2023-ek} explored using the Rust for Linux Project to implement a
-native UDP driver in Rust in order to explore the performance the Rust programming language. The
+native UDP driver in Rust in order to explore the performance of the Rust programming language. The
 authors were able to get a basic driver working with performance only slightly slower than C using
-the Rust for Linux (RFL) project. The RFL project is still to immature to get a full driver up and
-running but is at a stage were we researchers can start experimenting with different approaches.
+the Rust for Linux (RFL) project. The RFL project is still too immature to get a full driver up and
+running, but it is at a stage where researchers can start experimenting with different approaches.
 
@@
 programming language. The non-encapsulated interface Rust component which was almost identical to
-the original C component only introduced a 0.7\% overhead. The encapsulated Rust component on the
-other hand added a 3\% performance overhead. Another important aspect of the kernel is size, we must
+the original C component, only introduced a 0.7\% overhead. The encapsulated Rust component, on the
+other hand, added a 3\% performance overhead. Another important aspect of the kernel is size, and we must
 be careful to keep the size of the Rust component as close as we can to the C component so we can
-still run on all the same hardware. The authors found that there was only a 0.06\% size increase
+still run on the same hardware. The authors found that there was only a 0.06\% size increase
 when compared to the original C implementation.
@@
 
-\subsection{RQ3 What are the major challenges, limitations, and lessons learned when
-      developing operating system kernels in Rust, particularly in comparison to other languages?}
+\subsection{RQ3: Challenges and Limitations}
 
@@
             \item Deeply ingrained monomorphization which increased the size of the rust binaries
-            \item Compiler optimizations are not as mature as some C based compilers thus increasing binary size
+            \item Compiler optimizations are not as mature as some C based compilers, thus increasing binary size
             \item Hidden data structures and data
@@
             \item Kernel interfaces, while designed for extensibility, are not designed for type safety
-            \item Hybrid Code Flow. The rust compiler can not track ownership when switching between modules written in C and Rust
+            \item Hybrid Code Flow. The Rust compiler can not track ownership when switching between modules written in C and Rust
         \end{itemize}
@@
 
-A primary concern for programming languages that are used for operating system development are the
+A primary concern for programming languages that are used for operating system development is the
 size of the binary especially in an embedded environment.  Ayers et al.~\cite{Ayers2022-sf} focused
-on reducing size of binaries produced by the rust compiler while working with Tock OS. They were
+on reducing the size of binaries produced by the rust compiler while working with Tock OS. They were
 able to identify several  causes of binary growth that are specific to the Rust programming language
-and have the following 5 recommendations when using Rust in a size constrained environment:
+and have the following 5 recommendations when using Rust in a size-constrained environment:
 \begin{itemize}
@@
 Burtsev et al.~\cite{Burtsev2021-mh} explored what is missing in the Rust programming language to
-help solve the isolation problem in an operating system. Currently Rust lacks the ability to express
+help solve the isolation problem in an operating system. Currently, Rust lacks the ability to express
 isolation in the heap without external support. For example, the RedLeaf experiential operating
-system which is written in Rust relied on an complex interface definition language (IDL) to enforce
+system, which was written in Rust, relied on a complex interface definition language (IDL) to enforce
 isolated heaps. The paper enumerates several properties of the Rust Language that could help with
@@
 
+\subsection{RQ4: Lessons Learned}
 Klimt et al.~\cite{Klimt2023-ob} details the lessons learned and challenges when implementing
 Theseus. First they found that it is impossible to write a complete operating system in 100\% safe
-Rust. For example when writing a memory management system raw pointers must be used in order to
-modify the hardware. Despite the limitations imposed by the hardware the authors describe how
-Theseus leveraged intralingual design to maximize the compilers role in enforcing correctness. By
-leveraging Rusts type system and borrow checker memory safety and correct ownership transfer can be
+Rust. For example, when writing a memory management system, raw pointers must be used to
+modify the hardware. Despite the limitations imposed by the hardware, the authors describe how
+Theseus leveraged intralingual design to maximize the compiler's role in enforcing correctness. By
+leveraging Rust's type system and borrow checker memory safety, correct ownership transfer can be
 achieved at a higher level than what could be done in C. The authors also detailed some of the
-limitations of intralingual design such as not being as expressive as many other formal verification
-techniques due to the limited invariant that can be enforced by the type system. One of the most
+limitations of intralingual design, such as not being as expressive as many other formal verification
+techniques due to the limited invariance that the type system can enforce. One of the most
 important lessons learned was the insight that a linear type system itself cannot guarantee
-uniqueness of the resource represented, such when a memory resources may overlap. The authors
+the uniqueness of the resource represented, such as when a memory resource may overlap. The authors
 introduce the idea of using a hybrid approach of verification where they leverage both the linear
@@
 systems. How do we ensure that ownership of memory is sound in the operating system itself? In
-operating systems written in C the kernel provides the ownership root to applications running on
-top, finding a new ownership root is a open research question with regards to using Rust for
+operating systems written in C, the kernel provides the ownership root to applications running on
+top. Finding a new ownership root is an open research question with regard to using Rust for
 operating system development.
@@
 
-In conducting this systematic literature review on the use of Rust for operating system development,
-several potential threats to validity must be acknowledged. These threats are categorized into
-internal, external, and conclusion validity, each presenting distinct challenges to the robustness
-and reliability of our findings.
+% In conducting this systematic literature review on the use of Rust for operating system development,
+% several potential threats to validity must be acknowledged. These threats are categorized into
+% internal, external, and conclusion validity, each presenting distinct challenges to the robustness
+% and reliability of our findings.
 
@@
 towards certain types of research, such as those reporting successful integration while
-underrepresenting studies detailing challenges and failures. Compared to C Rust is a brand new
+underrepresenting studies detailing challenges and failures. Compared to C, Rust is a brand new
 language that is still finding its way in the world.
 
-Internal Validity: Internal validity refers to the degree to which the review accurately establishes
+\textit{\textbf{Internal Threats.}} Internal validity refers to the degree to which the review accurately establishes
 causal relationships. Potential threats include:
@@
 
-External Validity: External validity pertains to the generalizability of the review's findings
+\textit{\textbf{External Threats.}} External validity pertains to the generalizability of the review's findings
 beyond the specific context of the reviewed studies. Threats to external validity include:
@@
 
-Conclusion Validity: Conclusion validity involves the degree to which the conclusions drawn from the
+\textbf{\textit{Conclusion Threats}.} Conclusion validity involves the degree to which the conclusions drawn from the
 review are credible and supported by the data. Potential threats include:
@@
 
-\section{Conclusion}
+\section{Discussion \& Conclusion}
 
@@
 development, promising to address long-standing issues related to memory safety and system
-reliability. This systematic literature review has synthesized the current state of research and
-practice concerning the use of Rust in multiple types of kernels including the venerable Linux
+reliability. We synthesized the current state of research and
+practice concerning the use of Rust in multiple types of kernels, including the venerable Linux
 kernel, highlighting both the progress made and the challenges that remain.
@@
 
-However, the adoption of Rust for kernel development is not without its challenges. Truly the
+However, the adoption of Rust for kernel development is not without its challenges. Truly, the
 biggest challenge is existing codebases and mountains of legacy code. With extensive use of C and
-reliance on specific C idioms and low-level programming techniques researchers have substantial
+reliance on specific C idioms and low-level programming techniques, researchers have substantial
 integration hurdles to overcome. Any efforts to rewrite substantial portions of an existing kernel
-such as Linux in Rust are constrained by the need for interoperability with existing C code and the
+such as Linux in Rust, are constrained by the need for interoperability with existing C code and the
 necessity to maintain the kernel's performance characteristics. Furthermore, the operating system
-development community must reach consensus on Rust's role and ensure that sufficient tooling,
+development community must reach a consensus on Rust's role and ensure that sufficient tooling,
 documentation, and support are available for developers.

### 2024-06-05 paper.tex
Update on Overleaf.

@@
 For our research methodology, we followed the Kitchenham and Charters~\cite{Stuart2007-cc} methodology to conduct SLR in software engineering.
-We divided our review into three discrete phases, planning the review, conducting the review, and
-reporting the review results. The following sections detail our review process.
+We divided our review into three discrete phases, planning the review, conducting the review, and reporting the review results.
+The following sections detail our review process.
 
@@
 software developers to use, Rust is one of the few languages that is feasible to use when developing
-operating system kernels due to its lack of runtime and garbage collector. Even the Linux kernel is
+operating system kernels due to its lack of runtime and garbage collector~\cite{Ojeda2021-ul}. Even the Linux kernel is
 in the early stages of adding real support for
@@
 safety guarantees, its relatively recent adoption in systems programming, and the inherent
-complexities of kernel development. This section discusses these challenges, limitations, and the
+complexities of kernel development~\cite{Ojeda2021-ul}. This section discusses these challenges, limitations, and the
 lessons learned from various projects and research efforts. We summarize our findings in
@@
 \subsection{RQ4: Lessons Learned}
+
 Klimt et al.~\cite{Klimt2023-ob} details the lessons learned and challenges when implementing
@@
 type system and an SMT solver. The authors also explore the bootstrapping problem with rust
-systems. How do we ensure that ownership of memory is sound in the operating system itself? In
-operating systems written in C, the kernel provides the ownership root to applications running on
+systems. In operating systems written in C, the kernel provides the ownership root to applications running on
 top. Finding a new ownership root is an open research question with regard to using Rust for
@@
 % and reliability of our findings.
-
-Definition Ambiguity: The term "Kernel development in Rust" encompasses a wide range of activities,
-from minor module development to complete subsystem rewrites. Variations in how researchers and
-developers interpret this integration could lead to inconsistent findings. Selection Criteria: The
-inclusion and exclusion criteria for selecting relevant studies might inadvertently bias the review
-towards certain types of research, such as those reporting successful integration while
-underrepresenting studies detailing challenges and failures. Compared to C, Rust is a brand new
-language that is still finding its way in the world.
 
@@
     reliability of their findings.
+  \item  Selection Criteria: The inclusion and exclusion criteria for selecting relevant studies might inadvertently bias the review         towards certain types of research, such as those reporting successful integration while
+    underrepresenting studies detailing challenges and failures.
 \end{itemize}
@@
     and changes in both domains can alter the landscape significantly.
+    \item Definition Ambiguity: The term "Kernel development in Rust" encompasses a wide range of activities,
+    from minor module development to complete subsystem rewrites. Variations in how researchers and
+    developers interpret this integration could lead to inconsistent findings.
 \end{itemize}

### 2024-06-06 paper.tex
Update on Overleaf.

@@
 
-Comparing performance, throughput and latency between different systems is very difficult. For
-example, several of the operating systems written in Rust, such as Tock, are a complete rethinking of
-how an operating system is designed, and Rust for Linux (RFL) is still relatively new and has not had
-the benefit of decades of incremental performance improvements like components that are written in
-C. Additionally, direct comparisons between a Monolithic kernel, Micro-kernel, and Unikernel, or
+Comparing performance, throughput and latency between different systems with different architectures is very difficult. For
+example, several of the operating systems written in Rust are a complete rethinking of
+how an operating system is designed~\cite{Boos2020-zh} while others run entirely in kernel space~\cite{Lankes2019-cm}.
+Additionally, direct comparisons between a Monolithic kernel, Micro-kernel, and Unikernel, or
 comparisons between a Real Time Operation System (RTOS) and a General Purpose Operating System, are
-not directly meaningful due to the vastly different goals of each system. Therefore, we will focus
+not directly meaningful due to the vastly different goals of each system and the overhead that is imposed by the hardware~\cite{{Arnott2012-fl}. Therefore, we will focus
 more on the overall efforts that are specific to the Rust programming language and the challenges

### 2024-06-06 paper.tex
Update on Overleaf.

@@
 
-Unikernels present an ideal surface for prototyping Rust. Unikernels are designed to do one thing
+Unikernels present an ideal surface for prototyping Rust~\cite{Madhavapeddy2014-zw}. Unikernels are designed to do one thing
 and one thing only, thus limiting the complexity that a general-purpose operating system has to
@@
 comparisons between a Real Time Operation System (RTOS) and a General Purpose Operating System, are
-not directly meaningful due to the vastly different goals of each system and the overhead that is imposed by the hardware~\cite{{Arnott2012-fl}. Therefore, we will focus
-more on the overall efforts that are specific to the Rust programming language and the challenges
+not directly meaningful due to the vastly different goals of each system and the overhead that is imposed by the hardware~\cite{Arnott2012-fl}.
+Therefore, we will focus more on the overall efforts that are specific to the Rust programming language and the challenges
 presented. We summarize our findings in table~\ref{tab:RQ2}.
 
-Culic et al.~\cite{Culic2022-bk} looked at latency issues in the Tock OS. Tock is a new operating
-system that is designed to run on embedded systems but does not provide Real-time capabilities. The
+Culic et al.~\cite{Culic2022-bk} looked at latency issues in Tock. Tock is a new operating
+system written in Rust that is designed to run on embedded systems but does not provide Real-time capabilities. The
 authors attempted to add real-time capabilities by integrating eBPF into the Tock kernel to improve

### 2024-06-06 paper.tex
Update on Overleaf.

@@
 We divided our review into three discrete phases, planning the review, conducting the review, and reporting the review results.
-The following sections detail our review process.
+The following sections detail our review process and are diagrammed in figure~\ref{fig:process}.
+
+ \begin{figure*}
+    \centering
+     \includegraphics[width=7in]{process-diagram.pdf}
+     \caption{Process Diagram for Paper selection}
+     \label{fig:process}
+ \end{figure*}
 
@@
 software developers to use, Rust is one of the few languages that is feasible to use when developing
-operating system kernels due to its lack of runtime and garbage collector~\cite{Ojeda2021-ul}. Even the Linux kernel is
+operating system kernels due to its lack of runtime and garbage collector~\cite{Ojeda2021-ul}. The Linux kernel is currently
 in the early stages of adding real support for
@@
     \item \textbf{RQ1:} What are the existing approaches for implementing operating system kernels in Rust?
+
+    All though Rust in the Linux kernel is very recent we would like to see what approaches researches have explored in any operating system.
+Experiential and research kernels can provide new ideas that can be integrated into the Linux kernel.
     \item \textbf{RQ2:} What are the performance implications of using Rust for operating system kernel development?
+
+    The C programming language has the advantage of 40 years of improvements and lessons learned regarding how to write fast code.
+Rust being a new language still has many unexplored areas with how to write fast code especially in kernel space.
       %in terms of throughput, latency, and resource utilization?
@@
       developing operating system kernels in Rust?% particularly in comparison to other languages?
+
+      Kernel space presents unique challenges for the software engineer using Rust.
+Rust was not designed to implement kernels and is more restrictive than C.
+Thus, this research question is looking for any stumbling blocks that have already been found.
     \item \textbf{RQ4:} What are the lessons learned when developing operating systems kernels in Rust?
+
+    After the work has been completed we want to find all the research that reflects on what went well and what didn't.
 \end{itemize}
@@
 completed we were left with 26 potential papers that we needed an in-depth reading with an
-additional 2 papers were found during forward and backward snowballing.
+additional 6 papers were found during forward and backward snowballing.
 
@@
 
-In this section, we present our findings. We address each research question from \textbf{RQ1} - \textbf{RQ3}.
+In this section, we present our findings. We address each research question from \textbf{RQ1} - \textbf{RQ4}.
 
@@
 where the operating system architecture is redesigned and built from the ground up in order to fully
-leverage the Rust programming language. The second approach is the incremental integration of Rust
+leverage the Rust programming language~\cite{Culic2022-bk, Boos2020-zh}. The second approach is the incremental integration of Rust
 into existing kernels, such as the Linux
-kernel~\cite{The_kernel_development_community_undated-iw}. This methodology involves gradually
+kernel~\cite{The_kernel_development_community_undated-iw, Miller2021-pg, Oikawa2023-ms}. This methodology involves gradually
 rewriting or augmenting specific components of the kernel with Rust while maintaining compatibility
-with the existing C codebase. Table~\ref{tab:RQ1} summarizes our findings regarding what existing
+with the existing C codebase. Finally, researches are porting existing C kernels over to Rust and then evaluating the differences. Table~\ref{tab:RQ1} summarizes our findings regarding what existing
 approaches researchers are using to integrate rust into the kernel.
@@
 of matching the hardware.
+Li and Sato~\cite{Li2024-yb} explored using Rust to write an exokernel named W-Kernel.
+The authors proposed a novel architecture for an OS by embedding a WebAssembly(WASM) runtime into the kernel that can run programs written in any language that can compile down to WASM.
 
-While single address space operating systems present a unique opportunity to push more
-responsibility into the compiler a more traditional micro-kernel implementation was taken by Chen X
-et al.~\cite{Chen2023-wb} and Liang et al.~\cite{Liang2021-bo}. By pushing most kernel functionality
+For the second approach researches take a more  traditional micro-kernel implementation~\cite{Chen2023-wb, Liang2021-bo}. By pushing most kernel functionality
 to user space, the authors limited the surface area that needs to be proved. Chen X et al.~\cite{Chen2023-wb} took
@@
 
-Unikernels present an ideal surface for prototyping Rust~\cite{Madhavapeddy2014-zw}. Unikernels are designed to do one thing
+Finally researchers have taken existing Unikernels written in C and ported them to to Rust. Unikernels are designed to do one thing
 and one thing only, thus limiting the complexity that a general-purpose operating system has to
-deal with. This reduced complexity was explored by Lankes et al.~\cite{Lankes2019-cm}. The authors
-detailed their experiences with rewriting an existing Unikernel, HermitCore, in Rust. The new
-Unikernel named RustyHermit consists of only 3.27\% unsafe Rust, with the rest
-of the code base consisting of safe Rust, dramatically decreasing the surface area where memory errors
-could originate.
-
-Finally, Li and Sato~\cite{Li2024-yb} explored using Rust to write an exokernel named W-Kernel. The authors
-proposed a novel architecture for an OS by embedding a WebAssembly(WASM) runtime into the kernel that can run programs written in any language that can compile down to WASM.
+deal with~\cite{Madhavapeddy2014-zw}. This reduced complexity allowed  Lankes et al.~\cite{Lankes2019-cm} to take HermitCore and rewrite it in Rust. The new
+Unikernel named RustyHermit consists of only 3.27\% unsafe Rust, with the rest of the code base consisting of safe Rust, dramatically decreasing the surface area where memory errors could originate.
 

### 2024-07-25 paper.tex
Update on Overleaf.

@@
-%\documentclass[sigconf]{acmart}
-\documentclass[sigconf,review,anonymous]{acmart}
+\documentclass[sigconf]{acmart}
+%\documentclass[sigconf,review,anonymous]{acmart}
 \usepackage{listings, listings-rust}
@@
 
-%% \author{Shane Panter}
-%%     \affiliation{%
-%%     \institution{Boise State University}
-%%     \city{Boise}
-%%     \state{Idaho}
-%%     \country{USA}}
-%%     \email{shanepanter@boisestate.edu}
+\author{Shane K. Panter}
+    \affiliation{%
+    \institution{Department of Computer Science}
+    \institution{Boise State University}
+    \city{Boise}
+    \state{Idaho}
+    \country{USA}}
+    \email{shanepanter@boisestate.edu}
 
-%% \author{Nasir U. Eisty}
-%%     \affiliation{%
-%%     \institution{Boise State University}
-%%     \city{Boise}
-%%     \state{Idaho}
-%%     \country{USA}}
-%%     \email{nasireisty@boisestate.edu}
+\author{Nasir U. Eisty}
+    \affiliation{%
+    \institution{Department of Computer Science}
+    \institution{Boise State University}
+    \city{Boise}
+    \state{Idaho}
+    \country{USA}}
+    \email{nasireisty@boisestate.edu}
 
@@
 
-\received{30 May 2024}
-\received[revised]{TBD}
-\received[accepted]{TBD}
+% \received{30 May 2024}
+% \received[revised]{TBD}
+\received[accepted]{22 July 2024}
 
@@
 
-The 1995 movie Hackers~\cite{Wikipedia_contributors2024-zr} prescient predictions regarding the ease
+The 1995 movie Hackers prescient predictions regarding the ease
 of breaking into computing systems in cyberspace have come to fruition. The White House Office of
@@
 
-\section{Research Methodology}
-
-For our research methodology, we followed the Kitchenham and Charters~\cite{Stuart2007-cc} methodology to conduct SLR in software engineering.
-We divided our review into three discrete phases, planning the review, conducting the review, and reporting the review results.
-The following sections detail our review process and are diagrammed in figure~\ref{fig:process}.
-
  \begin{figure*}
@@
  \end{figure*}
+
+\section{Research Methodology}
+
+For our research methodology, we followed the Kitchenham and Charters~\cite{Stuart2007-cc} methodology to conduct SLR in software engineering.
+We divided our review into three discrete phases, planning the review, conducting the review, and reporting the review results.
+The following sections detail our review process and are diagrammed in figure~\ref{fig:process}.
 
@@
 
-\subsubsection{\textbf{Research Questions}}
+\subsection{Research Questions}
 \label{sec:researchQuestions}
 
-We have defined the following research questions to drive this study.
+We have defined the following research questions:
 
 \begin{itemize}
-    \item \textbf{RQ1:} What are the existing approaches for implementing operating system kernels in Rust?
+    \item \textbf{RQ1: What are the existing approaches for implementing operating system kernels in Rust?}
 
-    All though Rust in the Linux kernel is very recent we would like to see what approaches researches have explored in any operating system.
+    Although Rust in the Linux kernel is recent, we want to see what approaches researchers have explored in any operating system.
 Experiential and research kernels can provide new ideas that can be integrated into the Linux kernel.
-    \item \textbf{RQ2:} What are the performance implications of using Rust for operating system kernel development?
+    \item \textbf{RQ2: What are the performance implications of using Rust for operating system kernel development?}
 
     The C programming language has the advantage of 40 years of improvements and lessons learned regarding how to write fast code.
-Rust being a new language still has many unexplored areas with how to write fast code especially in kernel space.
-      %in terms of throughput, latency, and resource utilization?
-    \item \textbf{RQ3:} What are the major challenges and limitations when
-      developing operating system kernels in Rust?% particularly in comparison to other languages?
+Rust, being a new language, still has many unexplored areas regarding how to write fast code, especially in kernel space.
+Therefore, we want to explore the performance implications of Rust for kernel development in terms of throughput, latency, and resource utilization.
+    \item \textbf{RQ3: What are the major challenges and limitations when developing operating system kernels in Rust?}% particularly in comparison to other languages?
 
@@
 Rust was not designed to implement kernels and is more restrictive than C.
-Thus, this research question is looking for any stumbling blocks that have already been found.
-    \item \textbf{RQ4:} What are the lessons learned when developing operating systems kernels in Rust?
+Thus, we want to identify if there are any stumbling blocks that have already been found when using Rust in kernel development.
+    \item \textbf{RQ4: What are the lessons learned when developing operating systems kernels in Rust?}
 
-    After the work has been completed we want to find all the research that reflects on what went well and what didn't.
+    Besides technical challenges and limitations, developers' perceptions are very important in software development.
+Therefore, through this research question, we want to find developers' reflections on what went well and what didn't go well while developing kernels in Rust.
 \end{itemize}
@@
 
-We employed a robust multi-step search strategy across three databases: ACM Digital Library,
-IEEE Xplore, and Scopus to find all the current research. 
+We employed a robust multi-step search strategy across four databases: ACM Digital Library,
+IEEE Xplore, Scopus, and Springer Link to find all the current research. 
 %These databases cover a wide range of current software engineering and computer science literature. 
@@
 
-We searched all three databases with the keywords listed in table~\ref{tab:keywords} between January
-1, 2019 and April 1, 2024. While both ACM and IEEE had mutually exclusive results, Scopus had 44
-duplicates that needed to be removed. We then used Google Scholar for forward and backward
-snowballing. We choose to look at only the previous 5 years of research in order to capture the
-bleeding-edge research that is currently being done.
+We searched all four databases with the keywords listed in table~\ref{tab:keywords} between January
+1, 2019 and April 1, 2024. We choose to look at only the previous 5 years of research in order to capture the
+bleeding-edge research that is currently being done. While both ACM and IEEE had mutually exclusive results, Scopus had 44
+duplicates that needed to be removed and Springer Link didn't have any new papers that were not also in ACM or IEEE. We then used Google Scholar to do forward and backward snowballing.
 
@@
  & 87 \\
+ Springer Link & Operating System Rust Kernel (Conference Paper, Article, Research article) Subdiscipline: Software engineering/programming and operating systems & 29 \\
  \hline
@@
 
+\begin{table*}
+    \begin{tabular}{||l|l|l||}
+    \hline
+    Approach & Papers & Operating System in Rust\\
+    \hline\hline
+    Monolithic  & \cite{The_kernel_development_community_undated-iw}, \cite{Li2019-ru}, \cite{Miller2021-pg}, \cite{Oikawa2023-ms} & \href{https://docs.kernel.org/rust/}{Linux kernel v6.1+}\\
+    Micro-kernel & \cite{Chen2023-wb},\cite{Liang2021-bo}, \cite{Liu2024-xe}, \cite{Narayanan2020-gs}, \cite{Narayanan2019-fd} & Atmosphere, Redox, Redleaf\\
+    Embedded & \cite{Culic2022-bk}, \cite{Vishnunaryan2022-yd} & \href{https://github.com/tock/tock}{Tock}, \href{https://hubris.oxide.computer/}{Hubris}, \href{https://www.drone-os.com/}{Drone}, \href{https://bern-rtos.org/}{Bern}, HarSaRK \\
+    Unikernel & \cite{Lankes2019-cm},  \cite{Boos2020-zh}, \cite{Ijaz2023-da}, \cite{Sung2020-bb}  & RustyHermit, Theseus \\
+    Exokernel & \cite{Li2024-yb} & W-Kernel \\
+    \hline
+  \end{tabular}
+  \caption{Approaches and Methodologies for Rust in the Kernel}
+    \label{tab:RQ1}
+\end{table*}
+
 Following the guidelines outlined by Kitchenham and Charters~\cite{Stuart2007-cc} we set the
 inclusion and exclusion criteria based on our research questions outlined in
-section~\ref{sec:researchQuestions}. We only consider papers that are written in English, published
-in conferences, journals, workshops, and within the time frame of 2019 and 2024. The published papers should
+section~\ref{sec:researchQuestions}. We only considered papers that are written in English and published
+in conferences, journals, and workshops. The published papers should
 describe using the Rust programming language for either developing a new kernel, extending an
-existing kernel, or authoring drivers. We include any type of kernel architecture, including
+existing kernel, or authoring drivers. We included any type of kernel architecture, including
 monolithic kernel, microkernel, or unikernel in both the embedded and non-embedded space, as long as
 the paper is using the Rust programming language in some way. Papers that describe solutions that
-reside 100\% in user space will be excluded. Early efforts to use Rust in kernel development
+reside 100\% in user space or papers that relied too heavily on a component written in C were excluded. Early efforts to use Rust in kernel development
 were impacted by changes and updates in both the Rust language and compiler and relied on unstable
-features, which are thus excluded from this review.
+features, which were thus excluded from this review.
 
@@
 one record and then started our first pass, which consisted of reading the title and abstract and
-either marking the paper as \textit{include} or \textit{exclude}. Once the initial screening was
-completed we were left with 26 potential papers that we needed an in-depth reading with an
-additional 6 papers were found during forward and backward snowballing.
+either marking the paper as \textit{include} or \textit{exclude}.
+Once we completed the initial screening, we were left with 26 papers that needed an in-depth reading and analysis.
+We found an additional 6 papers during forward and backward snowballing that were also reviewed, bringing the total number of papers to 32.
+The first author manually reviewed all 32 papers, built the first dataset, and conducted the analysis.
+The second author then verified all the data and results.
+Since the results were straightforward and came from a comparatively small number of papers, there were no discrepancies or missing results to solve.
 
@@
 induced by interacting with other services. That state change can eventually cause a system crash or
-system instability at some point in the future making it very difficult to track down the problem
+system instability at some point in the future, making it very difficult to track down the problem
 due to the separation between when the problem occurred and when it was observed. Theseus OS is an
@@
 Li and Sato~\cite{Li2024-yb} explored using Rust to write an exokernel named W-Kernel.
-The authors proposed a novel architecture for an OS by embedding a WebAssembly(WASM) runtime into the kernel that can run programs written in any language that can compile down to WASM.
+The authors proposed a novel architecture for an OS by embedding a WebAssembly (WASM) runtime into the kernel that can run programs written in any language that can compile down to WASM.
 
-For the second approach researches take a more  traditional micro-kernel implementation~\cite{Chen2023-wb, Liang2021-bo}. By pushing most kernel functionality
+For the second approach, researchers take a more traditional micro-kernel implementation~\cite{Chen2023-wb, Liang2021-bo}. By pushing most kernel functionality
 to user space, the authors limited the surface area that needs to be proved. Chen X et al.~\cite{Chen2023-wb} took
 their implementation one step further by formally verifying their new micro-kernel named Atmosphere
-by using both the liner type system of Rust in combination with a SMT solver. They were
+by using both the liner type system of Rust in combination with an SMT solver. They were
 able to get a 7.5:1 proof-to-code ratio, which is higher than other formerly verified micro-kernels
@@
 
-Finally researchers have taken existing Unikernels written in C and ported them to to Rust. Unikernels are designed to do one thing
+Finally, researchers have taken existing Unikernels written in C and ported them to Rust. Unikernels are designed to do one thing
 and one thing only, thus limiting the complexity that a general-purpose operating system has to
@@
 
-\begin{table*}
+\subsection{RQ2: Performance Implications}
+
+\begin{table}
     \begin{tabular}{||l|l|l||}
     \hline
-    Approach & Papers & Operating System in Rust\\
+    No. & Implication & Studies that Reported the challenge\\
     \hline\hline
-    Monolithic  & \cite{The_kernel_development_community_undated-iw}, \cite{Li2019-ru}, \cite{Miller2021-pg}, \cite{Oikawa2023-ms} & \href{https://docs.kernel.org/rust/}{Linux kernel v6.1+}\\
-    Micro-kernel & \cite{Chen2023-wb},\cite{Liang2021-bo}, \cite{Liu2024-xe}, \cite{Narayanan2020-gs}, \cite{Narayanan2019-fd} & Atmosphere, Redox, Redleaf\\
-    Embedded & \cite{Culic2022-bk}, \cite{Vishnunaryan2022-yd} & \href{https://github.com/tock/tock}{Tock}, \href{https://hubris.oxide.computer/}{Hubris}, \href{https://www.drone-os.com/}{Drone}, \href{https://bern-rtos.org/}{Bern}, HarSaRK \\
-    Unikernel & \cite{Lankes2019-cm},  \cite{Boos2020-zh}, \cite{Ijaz2023-da}, \cite{Sung2020-bb}  & RustyHermit, Theseus \\
-    Exokernel & \cite{Li2024-yb} & W-Kernel \\
+    1 & Performance & \cite{Gonzalez2023-ek}, \cite{Li2024-be}, \cite{Ma2023-ef}\\
+    2 & Throughput & \cite{Gonzalez2023-ek}\\
+    3 & Latency & \cite{Culic2022-bk} \\
     \hline
   \end{tabular}
-  \caption{Approaches and Methodologies for Rust in the Kernel}
-    \label{tab:RQ1}
-\end{table*}
-
-\subsection{RQ2: Performance Implications}
+  \caption{Performance Implications of Rust in the Kernel}
+    \label{tab:RQ2}
+\end{table}
 
@@
 Culic et al.~\cite{Culic2022-bk} looked at latency issues in Tock. Tock is a new operating
-system written in Rust that is designed to run on embedded systems but does not provide Real-time capabilities. The
+system written in Rust that is designed to run on embedded systems but does not provide real-time capabilities. The
 authors attempted to add real-time capabilities by integrating eBPF into the Tock kernel to improve
-the interrupt handler's response time. The Authors found that early work (still in the prototype
+the interrupt handler's response time. The authors found that early work (still in the prototype
 stage) lowers the response times of the system and interrupts response times 3x.
@@
 Li et al.~\cite{Li2024-be} explored the feasibility of using Rust in kernel space. The authors took
-an existing component, the Out of Memory (OOM) and implemented a replacement using the Rust
+an existing component, the Out of Memory (OOM), and implemented a replacement using the Rust
 programming language. The non-encapsulated interface Rust component which was almost identical to
 the original C component, only introduced a 0.7\% overhead. The encapsulated Rust component, on the
-other hand, added a 3\% performance overhead. Another important aspect of the kernel is size, and we must
-be careful to keep the size of the Rust component as close as we can to the C component so we can
-still run on the same hardware. The authors found that there was only a 0.06\% size increase
-when compared to the original C implementation.
-
-\begin{table}
-    \begin{tabular}{||l|l|l||}
-    \hline
-    No. & Implication & Studies that Reported the challenge\\
-    \hline\hline
-    1 & Performance & \cite{Gonzalez2023-ek}, \cite{Li2024-be}, \cite{Ma2023-ef}\\
-    2 & Throughput & \cite{Gonzalez2023-ek}\\
-    3 & Latency & \cite{Culic2022-bk} \\
-    \hline
-  \end{tabular}
-  \caption{Performance Implications of using Rust in the Kernel}
-    \label{tab:RQ2}
-\end{table}
+other hand, added a 3\% performance overhead.
 
@@
 
-\begin{table*}[htbp]
+\begin{table*}
     \begin{tabular}{p{1cm} p{5cm} p{10cm}}
@@
             \item Provide software-only stack guard with extensible probing interface
-            \item Develop zero-copy serialization of “plain- old” data structures
+            \item Develop zero-copy serialization of “plain-old” data structures
         \end{itemize}
@@
 A primary concern for programming languages that are used for operating system development is the
-size of the binary especially in an embedded environment.  Ayers et al.~\cite{Ayers2022-sf} focused
+size of the binary especially in an embedded environment, we must
+be careful to keep the size of the Rust component as close as we can to the C component so we can
+still run on the same hardware. Li et al.~\cite{Li2024-be} found that there was only a 0.06\% size increase when compared to the original C implementation.  Ayers et al.~\cite{Ayers2022-sf} focused
 on reducing the size of binaries produced by the rust compiler while working with Tock OS. They were
@@
 Klimt et al.~\cite{Klimt2023-ob} details the lessons learned and challenges when implementing
-Theseus. First they found that it is impossible to write a complete operating system in 100\% safe
+Theseus. First, they found that it is impossible to write a complete operating system in 100\% safe
 Rust. For example, when writing a memory management system, raw pointers must be used to
@@
 
-\textit{\textbf{Internal Threats.}} Internal validity refers to the degree to which the review accurately establishes
-causal relationships. Potential threats include:
+\textit{\textbf{Internal Threats.}} 
+%Internal validity refers to the degree to which the review accurately establishes
+%causal relationships.
+Potential internal threats include:
 
 \begin{itemize}
-  \item Publication Bias: Studies demonstrating positive outcomes of using Rust for kernel
-    development may be more likely to be published and cited, leading to an overestimation of its
-    benefits.
-  \item Quality of Studies: The methodological rigor of the included studies varies, and some may
+  %\item Publication Bias: Studies demonstrating positive outcomes of using Rust for kernel
+    %development may be more likely to be published and cited, leading to an overestimation of its
+    %benefits.
+  \item \textit{Quality of Studies:} The methodological rigor of the included studies varies, and some may
     suffer from design flaws or biases that were not adequately controlled for, thus affecting the
     reliability of their findings.
-  \item  Selection Criteria: The inclusion and exclusion criteria for selecting relevant studies might inadvertently bias the review         towards certain types of research, such as those reporting successful integration while
-    underrepresenting studies detailing challenges and failures.
+  \item  \textit{Selection Criteria:} The inclusion and exclusion criteria for selecting relevant studies might inadvertently bias the review towards certain types of research, such as those reporting successful integration while underrepresenting studies detailing challenges and failures.
 \end{itemize}
 
-\textit{\textbf{External Threats.}} External validity pertains to the generalizability of the review's findings
-beyond the specific context of the reviewed studies. Threats to external validity include:
+\textit{\textbf{External Threats.}} 
+%External validity pertains to the generalizability of the review's findings
+%beyond the specific context of the reviewed studies. 
+Threats to external validity include:
 
 \begin{itemize}
-  \item Context-Specific Findings: Many studies focus on specific kernel modules or use-cases, which
+  \item \textit{Context-Specific Findings:} Many studies focus on specific kernel modules or use-cases, which
     may not be representative of the broader kernel environment. The success of Rust in isolated
     components does not necessarily translate to the entire kernel.
-  \item Temporal Changes: The rapidly evolving nature of both the Rust programming language and the
+  \item \textit{Temporal Changes:} The rapidly evolving nature of both the Rust programming language and the
     Linux kernel means that findings from older studies may no longer be applicable, as improvements
     and changes in both domains can alter the landscape significantly.
-    \item Definition Ambiguity: The term "Kernel development in Rust" encompasses a wide range of activities,
+    \item \textit{Definition Ambiguity:} The term ``Kernel development in Rust'' encompasses a wide range of activities,
     from minor module development to complete subsystem rewrites. Variations in how researchers and
@@
 
-\textbf{\textit{Conclusion Threats}.} Conclusion validity involves the degree to which the conclusions drawn from the
-review are credible and supported by the data. Potential threats include:
+\textbf{\textit{Conclusion Threats}.}
+%Conclusion validity involves the degree to which the conclusions drawn from the
+%review are credible and supported by the data.
+Potential conclusion threats include:
 
 \begin{itemize}
-  \item Heterogeneity of Studies: The diverse methodologies, metrics, and contexts of the included studies
+  \item \textit{Heterogeneity of Studies:} The diverse methodologies, metrics, and contexts of the included studies
     can lead to challenges in synthesizing findings and drawing unified conclusions. For example, it
     is very difficult to directly compare a micro-kernel, Unikernel, and hybrid Kernel.
-  \item Reviewer Bias: Personal biases of the reviewers in interpreting data and making judgments
+  \item \textit{Reviewer Bias:} Personal biases of the reviewers in interpreting data and making judgments
     about study quality and relevance may skew the results
@@
 
-Efforts were made to minimize this through predefined criteria and multiple reviewers, but some
+We made efforts to minimize this through predefined criteria and multiple reviewers, but some
 degree of subjectivity is inevitable.  In addressing these threats, we employed rigorous methods for
 study selection, data extraction, and analysis, and we remain transparent about the
-limitations. Future reviews could mitigate these threats further by incorporating a broader range of
-sources, using more precise definitions, and employing meta-analytic techniques to quantitatively
-synthesize findings. Despite these threats, our review provides a valuable synthesis of the current
+limitations. 
+%Future studies could mitigate these threats further by incorporating a broader range of sources, using more precise definitions, and employing meta-analytic techniques to quantitatively synthesize findings.
+Despite these threats, our results provide a valuable synthesis of the current
 state of research on the use of Rust in kernel development, highlighting both its potential benefits

### 2024-08-30 paper.tex
Update on Overleaf.

@@
 \documentclass[sigconf]{acmart}
-%\documentclass[sigconf,review,anonymous]{acmart}
-\usepackage{listings, listings-rust}
-\usepackage{xcolor}
-\usepackage[raggedrightboxes]{ragged2e}
-
-\pagestyle{plain}
-
-\NewDocumentCommand{\codeword}{v}{%
-\texttt{\textcolor{blue}{#1}}
-}
-
-\lstset{language=Rust, keywordstyle={\bfseries \color{blue}}}
+% Camera Ready documentation
+% https://www.acm.org/binaries/content/assets/publications/taps/latex-best_practices-06-may-2020.pdf
 
 %% Rights management information.
-\setcopyright{acmlicensed}
 \copyrightyear{2024}
 \acmYear{2024}
-\acmDOI{XXXXXXX.XXXXXXX}
-%\acmSubmissionID{123-A56-BU3}
+\setcopyright{rightsretained}
+\acmConference[ESEM '24]{Proceedings of the 18th ACM / IEEE International Symposium on Empirical Software Engineering and Measurement}{October 24--25, 2024}{Barcelona, Spain}
+\acmBooktitle{Proceedings of the 18th ACM / IEEE International Symposium on Empirical Software Engineering and Measurement (ESEM '24), October 24--25, 2024, Barcelona, Spain}
+\acmDOI{10.1145/3674805.3690756}
+\acmISBN{979-8-4007-1047-6/24/10}
 
 \begin{document}
-%\title{Kernel Development Using Rust: A Systematic Literature Review}
 \title{Rusty Linux: Advances in Rust for Linux Kernel Development}
-
-%\author{Anonymous Authors}
 
@@
 
-%\renewcommand{\shortauthors}{Panter et al.}
-
 \begin{abstract}
 
-  \textit{\textbf{Context:}} The integration of Rust into kernel development is a transformative endeavor aimed
-  at enhancing system security and reliability by leveraging Rust's strong memory safety
-  guarantees. 
-  %This systematic literature review examines the current state of research and practice regarding the use of Rust in the kernel development.
-  \textit{\textbf{Objective:}} We aim to find the current advances in using Rust in Kernel
-  development to reduce the number of memory safety
-  vulnerabilities in one of the most critical pieces of software that underpins all modern
-  applications.
-  \textit{\textbf{Method:}} By analyzing a broad spectrum of studies, we identify the advantages Rust offers,
-  %such as reducing vulnerabilities like buffer overflows and use-after-free errors, which are
-  %prevalent in C-based kernel development. The review also 
-  highlight the challenges faced,
-  %including integration hurdles with existing C codebases, performance considerations, 
-  and emphasise the need for community consensus on Rust's adoption.
-  \textit{\textbf{Results:}} Our findings suggest that while the initial implementations of Rust in the kernel
-  show promising results in terms of safety and stability, significant challenges remain. These challenges
-  include achieving seamless interoperability with existing kernel components, maintaining
-  performance, and ensuring adequate support and tooling for developers.
-  \textit{\textbf{Conclusions:}} This study 
-  %provides a comprehensive overview of the
-  %potential and challenges of incorporating Rust into kernel development. It 
-  underscores the need for
-  continued research and practical implementation efforts to fully realize the benefits of Rust. By
-  addressing these challenges, the integration of Rust could mark a significant step forward in the
-  evolution of operating system development towards safer and more reliable systems.
+  \textit{\textbf{Context:}} The integration of Rust into kernel development is a transformative endeavor aimed at enhancing system security and reliability by leveraging Rust's strong memory safety guarantees.
+  \textit{\textbf{Objective:}} We aim to find the current advances in using Rust in Kerneldevelopment to reduce the number of memory safety vulnerabilities in one of the most critical pieces of software that underpins all modern applications.
+  \textit{\textbf{Method:}} By analyzing a broad spectrum of studies, we identify the advantages Rust offers, highlight the challenges faced, and emphasise the need for community consensus on Rust's adoption.
+  \textit{\textbf{Results:}} Our findings suggest that while the initial implementations of Rust in the kernel show promising results in terms of safety and stability, significant challenges remain.
+These challenges include achieving seamless interoperability with existing kernel components, maintaining performance, and ensuring adequate support and tooling for developers.
+  \textit{\textbf{Conclusions:}} This study  underscores the need for
+  continued research and practical implementation efforts to fully realize the benefits of Rust.
+By addressing these challenges, the integration of Rust could mark a significant step forward in the evolution of operating system development towards safer and more reliable systems.
 
@@
    <concept>
-<concept_id>10011007.10011006.10011008.10011024</concept_id>
-<concept_desc>Software and its engineering~Language features</concept_desc>
+       <concept_id>10002978.10003006.10003007</concept_id>
+       <concept_desc>Security and privacy~Operating systems security</concept_desc>
        <concept_significance>500</concept_significance>
@@
    <concept>
-<concept_id>10010520.10010521</concept_id>
-<concept_desc>Computer systems organization~Architectures</concept_desc>
+       <concept_id>10011007.10010940.10010941.10010949</concept_id>
+       <concept_desc>Software and its engineering~Operating systems</concept_desc>
+       <concept_significance>300</concept_significance>
+       </concept>
+   <concept>
+       <concept_id>10011007.10011006.10011008.10011024</concept_id>
+       <concept_desc>Software and its engineering~Language features</concept_desc>
        <concept_significance>500</concept_significance>
@@
 \ccsdesc[500]{General and reference~Surveys and overviews}
+\ccsdesc[500]{Security and privacy~Operating systems security}
+\ccsdesc[300]{Software and its engineering~Operating systems}
 \ccsdesc[500]{Software and its engineering~Language features}
-\ccsdesc[500]{Computer systems organization~Architectures}
 
@@
 
-% \received{30 May 2024}
-% \received[revised]{TBD}
 \received[accepted]{22 July 2024}
@@
      \caption{Process Diagram for Paper selection}
+     \Description{The process that we followed for selecting papers}
      \label{fig:process}
@@
  \hline\hline
- ACM  & (Abstract:("operating system" kernel linux OS) AND Abstract:(rust) )OR (Title:("operating
- system" OS kernel linux) AND Title:(rust))  & 30 \\
+ ACM  & (Abstract:("operating system" OR Kernel OR linux OR OS) AND Abstract:(Rust)) OR (Title:("operating system" OR OS OR kernel OR linux) AND Title:(Rust))  & 30 \\
  IEEE & ("All Metadata":"operating system" AND "All Metadata":rust) OR ("All Metadata":kernel AND
@@
 \begin{table*}
+    \caption{Approaches and Methodologies for Rust in the Kernel}
     \begin{tabular}{||l|l|l||}
@@
   \end{tabular}
-  \caption{Approaches and Methodologies for Rust in the Kernel}
   \label{tab:RQ1}
@@
 \begin{table}
+    \caption{Performance Implications of Rust in the Kernel}
     \begin{tabular}{||l|l|l||}
@@
   \end{tabular}
-  \caption{Performance Implications of Rust in the Kernel}
   \label{tab:RQ2}
@@
 \begin{table*}
+    \caption{Challenges Unique to the Rust Programming Language}
     \begin{tabular}{p{1cm} p{5cm} p{10cm}}
@@
   \end{tabular}
-  \caption{Challenges Unique to the Rust Programming Language}
   \label{tab:RQ3}
@@
 
-\balance
-
 \bibliographystyle{ACM-Reference-Format}
-\bibliography{slr-paper-memory-safety-rust.bib}
+\bibliography{paper.bib}
 

### 2024-08-30 paper.tex
Update on Overleaf.

@@
   \textit{\textbf{Context:}} The integration of Rust into kernel development is a transformative endeavor aimed at enhancing system security and reliability by leveraging Rust's strong memory safety guarantees.
-  \textit{\textbf{Objective:}} We aim to find the current advances in using Rust in Kerneldevelopment to reduce the number of memory safety vulnerabilities in one of the most critical pieces of software that underpins all modern applications.
+  \textit{\textbf{Objective:}} We aim to find the current advances in using Rust in Kernel development to reduce the number of memory safety vulnerabilities in one of the most critical pieces of software that underpins all modern applications.
   \textit{\textbf{Method:}} By analyzing a broad spectrum of studies, we identify the advantages Rust offers, highlight the challenges faced, and emphasise the need for community consensus on Rust's adoption.
@@
 
-For our research methodology, we followed the Kitchenham and Charters~\cite{Stuart2007-cc} methodology to conduct SLR in software engineering.
+For our research methodology, we followed the Kitchenham and Charters~\cite{Stuart2007-cc} methodology to conduct a SLR in software engineering.
 We divided our review into three discrete phases, planning the review, conducting the review, and reporting the review results.
@@
 
-Once we had our initial collection of papers from the databases, we merged all duplicate papers into
-one record and then started our first pass, which consisted of reading the title and abstract and
-either marking the paper as \textit{include} or \textit{exclude}.
-Once we completed the initial screening, we were left with 26 papers that needed an in-depth reading and analysis.
-We found an additional 6 papers during forward and backward snowballing that were also reviewed, bringing the total number of papers to 32.
-The first author manually reviewed all 32 papers, built the first dataset, and conducted the analysis.
+Once we had our initial collection of papers from the databases, we merged all duplicate papers into one record and then started our first pass, which consisted of reading the title and abstract and either marking the paper as \textit{include} or \textit{exclude}.
+Once we completed the initial screening, we were left with 22 papers that needed an in-depth reading and analysis.
+We found an additional 6 papers during forward and backward snowballing that were also reviewed, bringing the total number of papers to 28.
+The first author manually reviewed all 28 papers, built the first dataset, and conducted the analysis.
 The second author then verified all the data and results.

### 2024-08-30 paper.tex
Clean ups for camera ready

@@
 Therefore, we want to explore the performance implications of Rust for kernel development in terms of throughput, latency, and resource utilization.
-    \item \textbf{RQ3: What are the major challenges and limitations when developing operating system kernels in Rust?}% particularly in comparison to other languages?
+    \item \textbf{RQ3: What are the major challenges and limitations when developing operating system kernels in Rust?}
 
@@
 IEEE Xplore, Scopus, and Springer Link to find all the current research.
-%These databases cover a wide range of current software engineering and computer science literature. 
 We leveraged the advanced search features of all three databases to search the titles and abstracts for
@@
 
-Finally, researchers have taken existing Unikernels written in C and ported them to Rust. Unikernels are designed to do one thing
+Finally, researchers have taken existing unikernels written in C and ported them to Rust. Unikernels are designed to do one thing
 and one thing only, thus limiting the complexity that a general-purpose operating system has to
@@
 
-% In conducting this systematic literature review on the use of Rust for operating system development,
-% several potential threats to validity must be acknowledged. These threats are categorized into
-% internal, external, and conclusion validity, each presenting distinct challenges to the robustness
-% and reliability of our findings.
-
-\textit{\textbf{Internal Threats.}} 
-%Internal validity refers to the degree to which the review accurately establishes
-%causal relationships.
-Potential internal threats include:
+\textit{\textbf{Internal Threats.}} Potential internal threats include:
 
 \begin{itemize}
-  %\item Publication Bias: Studies demonstrating positive outcomes of using Rust for kernel
-    %development may be more likely to be published and cited, leading to an overestimation of its
-    %benefits.
   \item \textit{Quality of Studies:} The methodological rigor of the included studies varies, and some may
@@
 
-\textit{\textbf{External Threats.}} 
-%External validity pertains to the generalizability of the review's findings
-%beyond the specific context of the reviewed studies. 
-Threats to external validity include:
+\textit{\textbf{External Threats.}} Threats to external validity include:
 
@@
 
-\textbf{\textit{Conclusion Threats}.}
-%Conclusion validity involves the degree to which the conclusions drawn from the
-%review are credible and supported by the data.
-Potential conclusion threats include:
+\textbf{\textit{Conclusion Threats}.} Potential conclusion threats include:
 
@@
 study selection, data extraction, and analysis, and we remain transparent about the
-limitations. 
-%Future studies could mitigate these threats further by incorporating a broader range of sources, using more precise definitions, and employing meta-analytic techniques to quantitatively synthesize findings.
-Despite these threats, our results provide a valuable synthesis of the current
+limitations. Despite these threats, our results provide a valuable synthesis of the current
 state of research on the use of Rust in kernel development, highlighting both its potential benefits

### 2024-09-03 paper.tex
Minor edits for camera ready

@@
 able to get a 7.5:1 proof-to-code ratio, which is higher than other formerly verified micro-kernels
-SeL4 and CeriKOS both have proof code ratios of 19:1 and 20:1, respectively. While this
-approach was novel drivers running in userspace are still not formally verified and don't have to be
+SeL4 and CeriKOS both of which have proof code ratios of 19:1 and 20:1, respectively. While this
+approach was novel, drivers running in userspace are still not formally verified and don't have to be
 written in Rust.

### 2024-10-16 slides.tex
Preparing slides for conference

%%% Local Variables:
%%% mode: LaTeX
%%% TeX-master: t
%%% End:

\documentclass{beamer}
\usepackage[utf8]{inputenc}

\usetheme{Berkeley}

%\useoutertheme{shadow}
%\useinnertheme{rounded}

\definecolor{BSUblue}{RGB}{0, 51, 160} % (primary)
\definecolor{BSUorange}{RGB}{214,67, 9} % (secondary)

\setbeamercolor{palette primary}{bg=BSUblue,fg=white}
\setbeamercolor{palette secondary}{bg=BSUblue,fg=white}
\setbeamercolor{palette tertiary}{bg=BSUblue,fg=white}
\setbeamercolor{palette quaternary}{bg=BSUblue,fg=white}
\setbeamercolor{structure}{fg=BSUblue} % itemize, enumerate, etc
\setbeamercolor{section in toc}{fg=BSUblue} % TOC sections

% Override palette coloring with secondary
\setbeamercolor{subsection in head/foot}{bg=BSUorange,fg=white}

%------------------------------------------------------------
%Title page
\title[Rusty Linux] {Rusty Linux: Advances in Rust for Linux Kernel Development}

\author[]
{Shane K. Panter\inst{1} \and Nasir, Eisty\inst{2}}

\institute[BSU]
{
  \inst{1}%
  Clinical Assistant Professor\\
  Boise State University
  \and
  \inst{2}%
  Assistant Professor\\
  Boise State University
}

\date[ESEM 24]
{International Symposium on Empirical Software Engineering and Measurement, October 2024}

\logo{\includegraphics[height=1cm]{images/bsu-logo.eps}}

%End of title page configuration block
%------------------------------------------------------------

\begin{document}

\frame{\titlepage}

\section{Introduction}
\begin{frame}
  \frametitle{Introduction}
  \includegraphics[width=3in]{images/CCP-from-northwest.jpg}
  \begin{block}{Boise State University}
    The Computer Science Department is located in Beautiful downtown Boise Idaho, United States!
  \end{block}

\end{frame}

\section{Methodology}
\begin{frame}
  \frametitle{Process Diagram}
    \includegraphics[width=4in]{figures/process-diagram.pdf}
\end{frame}

\section{Results}

\end{document}

### 2024-10-17 slides.tex
Slides draft

@@
 \documentclass{beamer}
+
 \usepackage[utf8]{inputenc}
+\usepackage{graphicx}
+\usepackage{tikz}
+
+\usetikzlibrary {arrows.meta}
 
@@
   \frametitle{Introduction}
-  \includegraphics[width=3in]{images/CCP-from-northwest.jpg}
+  \begin{tikzpicture}
+    \node [inner sep=0pt,,outer sep=0pt,clip,rounded corners=0.5cm] (boise) at (0,0){
+      \includegraphics[width=0.8\textwidth]{images/boise-map.jpg}};
+
+    \node [inner sep=0pt,,outer sep=0pt,clip,rounded corners=0.5cm] (ccp) at (3,-1){
+      \includegraphics[width=0.45\textwidth]{images/CCP-from-northwest.jpg} };
+    \draw[line width=4pt,
+    arrows = {-Computer Modern Rightarrow[line cap=butt]}]       (ccp)   -- (-2.2,.8);
+
+  \end{tikzpicture}
+
   \begin{block}{Boise State University}
@@
 
+\subsection{Our Research}
+
+\begin{frame}
+  \frametitle{Our Research}
+  \begin{columns}
+    \begin{column}{0.5\textwidth}
+      We aim to find the current advances in using Rust in Kernel development to reduce the number of
+      memory safety vulnerabilities in one of the most critical pieces of software that underpins all
+      modern applications.
+    \end{column}
+    \begin{column}{0.5\textwidth}
+      \begin{figure}
+        \caption{A rusty computer\footnotemark[1]}
+        \includegraphics[width=.8\textwidth]{images/rusty.jpeg}\footnotemark[1]
+      \end{figure}
+    \end{column}
+  \end{columns}
+  \footnotetext[1]{AI Prompt: A rusty computer with a penguin next to it}
+  \href{https://github.com/shanep/esem-2024/blob/master/proofs/esem24-50.pdf}{\beamergotobutton{Paper
+    Link}}
+\end{frame}
+
 \section{Methodology}
+
+\subsection{Research Questions}
+\begin{frame}
+  \frametitle{Research Questions}
+  \begin{itemize}
+  \item<1-> \textbf{RQ1:} What are the existing approaches for implementing operating system kernels in Rust?
+  \item<2-> \textbf{RQ2:} What are the performance implications of using Rust for operating system kernel development?
+  \item<3-> \textbf{RQ3:} What are the major challenges and limitations when developing operating system kernels in Rust?
+  \item<4-> \textbf{RQ4:} What are the lessons learned when developing operating systems kernels in Rust?
+
+  \end{itemize}
+\end{frame}
+
+\subsection{Process Diagram}
 \begin{frame}
@@
 \end{frame}
-
 \section{Results}
 
+\begin{frame}
+  \frametitle{Results}
+  \begin{columns}
+    \begin{column}{0.5\textwidth}
+      Our findings!
+    \end{column}
+    \begin{column}{0.5\textwidth}
+      \begin{figure}
+        \caption{Super happy researcher!\footnotemark[1]}
+        \includegraphics[width=.8\textwidth]{images/results.jpeg}\footnotemark[1]
+      \end{figure}
+    \end{column}
+  \end{columns}
+  \footnotetext[1]{AI Prompt: scientist getting research results and is super happy in a cyberpunk
+    universe with lots of computers showing matrix code on them}
+\end{frame}
+
+\subsection{RQ1: Existing Approaches}
+\begin{frame}
+  \frametitle{RQ1}
+  \begin{table}
+    \caption{Approaches and Methodologies for Rust in the Kernel}
+    \begin{tabular}{||l|l|l||}
+      \hline
+      Approach & Papers & Operating System in Rust\\
+      \hline\hline
+      Monolithic   & 4 & Linux Kernel v6.1+\\
+      Micro-kernel & 5 & Atmosphere, Redox, Redleaf\\
+      Embedded     & 2 & Tock, Hubris, Drone, Bern, HarSaRK \\
+      Unikernel & 4 & RustyHermit, Theseus \\
+      Exokernel & 1 & W-Kernel \\
+      \hline
+    \end{tabular}
+    \label{tab:RQ1}
+  \end{table}
+
+\end{frame}
+
+\subsection{RQ2: Performance Implications}
+\begin{frame}
+  \frametitle{RQ2}
+  \begin{table}
+    \caption{Performance Implications of Rust in the Kernel}
+    \begin{tabular}{||l|l|l||}
+      \hline
+      No. & Implication & Studies that Reported the challenge\\
+      \hline\hline
+      1 & Performance & 3\\
+      2 & Throughput & 1\\
+      3 & Latency & 1 \\
+      \hline
+    \end{tabular}
+    \label{tab:RQ2}
+  \end{table}
+\end{frame}
+
+\subsection{RQ3: Challenges and Limitations}
+\begin{frame}
+  \frametitle{RQ3: Challenges and Limitations}
+  \begin{itemize}
+  \item<1-> Binary Size - Rust can produce larger binaries
+  \item<2-> Missing Features - Rust still evolving and adding features
+  \item<3-> Soundness - How to deal with raw memory
+  \item<4-> Panics - What happens when things go wrong?
+  \item<5-> C Interop - Specific to mixed language kernels
+  \end{itemize}
+\end{frame}
+
+\subsection{RQ4: Lessons Learned}
+
+\begin{frame}
+  \frametitle{RQ4: Lessons Learned}
+  \begin{itemize}
+  \item<1-> Impossible to use 100$\%$ rust
+  \item<2-> Rust is not as expressive as other formal verification techniques
+  \item<3-> Ownership root - An OS provides memory to rust so if the OS is itself written in rust
+    who is the root owner?
+  \end{itemize}
+
+\end{frame}
+
+\section{Conclusion}
+
+\begin{frame}
+  \frametitle{Conclusion}
+  \begin{itemize}
+  \item<1-> We are still in the early stages of figuring out who to do kernel dev in Rust
+  \item<2-> High potential for enhanced security and stability
+  \item<3-> Need to address integration issues (FFI)
+  \item<4-> Need to expand the body of empirical evidence on Rust's impact!
+  \end{itemize}
+
+\end{frame}
+
+\section{Questions?}
+
+\begin{frame}
+  \frametitle{Questions?}
+  \begin{columns}
+    \begin{column}{0.5\textwidth}
+      Questions?
+    \end{column}
+    \begin{column}{0.5\textwidth}
+      \begin{figure}
+        \caption{Happy People\footnotemark[1]}
+        \includegraphics[width=.8\textwidth]{images/questions.jpeg}\footnotemark[1]
+      \end{figure}
+    \end{column}
+  \end{columns}
+  \footnotetext[1]{AI Prompt: People attending a conference who all want to ask a question and are really excited!}
+\end{frame}
+
 \end{document}

### 2024-10-24 slides.tex
Removed a comma from my name
Minor updates to slides

@@
 
+%\useoutertheme{split}
 %\useoutertheme{shadow}
+%\useoutertheme{sidebar}
 %\useinnertheme{rounded}
+%\useinnertheme{inmargin}
 
@@
 \author[]
-{Shane K. Panter\inst{1} \and Nasir, Eisty\inst{2}}
+{Shane K. Panter\inst{1} \and Nasir Eisty\inst{2}}
 
@@
 
+\section{Why Rust}
+\begin{frame}
+  \frametitle{Why Rust}
+  \begin{itemize}
+  \item<1-> Low-level \textbf{control} like C and C++
+  \item<2-> Strong \textbf{safety} guarantees
+  \item<3-> \textbf{Modern}, functional paradigms
+  \item<4-> Industrial development and backing
+  \item<5-> No garbage collector needed! All checks are performed at compile time
+  \end{itemize}
+\end{frame}
+
 \section{Methodology}
@@
       Unikernel & 4 & RustyHermit, Theseus \\
-      Exokernel & 1 & W-Kernel \\
        \hline
@@
   \end{table}
+
+  \begin{itemize}
+  \item Performance issues - Caused by the safe $->$ unsafe transition layer
+  \item Throughput issues - Caused by immature and or missing bindings within the FFI layer
+  \item Latency issues - Caused by the interrupt layer written in Rust
+  \end{itemize}
+
 \end{frame}
@@
   \begin{itemize}
-  \item<1-> Binary Size - Rust can produce larger binaries
-  \item<2-> Missing Features - Rust still evolving and adding features
-  \item<3-> Soundness - How to deal with raw memory
+  \item<1-> Binary Size
+    \begin{itemize}
+    \item Rust can produce larger binaries
+    \item The same issue that C++ templates have!
+    \end{itemize}
+  \item<2-> Missing Features
+    \begin{itemize}
+    \item Rust still evolving and adding features
+    \item Makes it difficult to integrate into the classroom due to the rapid evolution
+    \end{itemize}
+  \item<3-> Soundness - How to deal with raw memory?
   \item<4-> Panics - What happens when things go wrong?
@@
     who is the root owner?
+    \begin{itemize}
+      \item Research is needed for hardware support (CHERI)
   \end{itemize}
@@
   \item<3-> Need to address integration issues (FFI)
-  \item<4-> Need to expand the body of empirical evidence on Rust's impact!
+  \item<4-> Need to expand the body of empirical evidence on Rust's impact! (Or more generally low
+    level memory safe languages)
   \end{itemize}
@@
 \end{document}
+
+%%% Local Variables:
+%%% mode: LaTeX
+%%% TeX-master: t
+%%% End:

### 2024-10-24 slides.tex
updated proofs dir

@@
       memory safety vulnerabilities in one of the most critical pieces of software that underpins all
-      modern applications.
+      modern applications (SLR).
     \end{column}
@@
     \begin{itemize}
-      \item Research is needed for hardware support (CHERI)
+    \item Open research question if this can even be done in software
+    \item Researchers looking at hardware support (CHERI)
+    \end{itemize}
   \end{itemize}

### 2024-10-24 slides.tex
Final updates to slides with ai gen pics

@@
 \subsection{Research Questions}
+
+\subsubsection{RQ1}
 \begin{frame}
-  \frametitle{Research Questions}
-  \begin{itemize}
-  \item<1-> \textbf{RQ1:} What are the existing approaches for implementing operating system kernels in Rust?
-  \item<2-> \textbf{RQ2:} What are the performance implications of using Rust for operating system kernel development?
-  \item<3-> \textbf{RQ3:} What are the major challenges and limitations when developing operating system kernels in Rust?
-  \item<4-> \textbf{RQ4:} What are the lessons learned when developing operating systems kernels in Rust?
+  \frametitle{RQ1}
+  \begin{columns}
+    \begin{column}{0.5\textwidth}
+      \textbf{RQ1:} What are the existing approaches for implementing operating system kernels in Rust?\footnotemark[2]
+    \end{column}
+    \begin{column}{0.5\textwidth}
+      \includegraphics[width=.8\textwidth]{images/rq1.jpeg}
+    \end{column}
+  \end{columns}
+  \footnotetext[2]{Fun side note: What happens if we put our research questions into an AI image generator?}
+\end{frame}
 
-  \end{itemize}
+\subsubsection{RQ2}
+\begin{frame}
+  \frametitle{RQ2}
+    \begin{columns}
+      \begin{column}{0.5\textwidth}
+        \textbf{RQ2:} What are the performance implications of using Rust for operating system kernel development?
+    \end{column}
+    \begin{column}{0.5\textwidth}
+      \includegraphics[width=.8\textwidth]{images/rq2.jpeg}
+    \end{column}
+  \end{columns}
+\end{frame}
+
+\subsubsection{RQ3}
+\begin{frame}
+  \frametitle{RQ3}
+  \begin{columns}
+    \begin{column}{0.5\textwidth}
+      \textbf{RQ3:} What are the major challenges and limitations when developing operating system kernels in Rust?
+    \end{column}
+    \begin{column}{0.5\textwidth}
+      \includegraphics[width=.8\textwidth]{images/rq3.jpeg}
+    \end{column}
+  \end{columns}
+\end{frame}
+
+\subsubsection{RQ4}
+\begin{frame}
+  \frametitle{RQ4}
+  \begin{columns}
+    \begin{column}{0.5\textwidth}
+      \textbf{RQ4:} What are the lessons learned when developing operating systems kernels in Rust?
+      \end{column}
+    \begin{column}{0.5\textwidth}
+      \includegraphics[width=.8\textwidth]{images/rq4.jpeg}
+    \end{column}
+  \end{columns}
 \end{frame}
@@
     \end{itemize}
-  \item<3-> Soundness - How to deal with raw memory?
+  \item<3-> Soundness - How to deal with raw memory without sacrificing safety?
   \item<4-> Panics - What happens when things go wrong?
@@
   \begin{itemize}
-  \item<1-> Impossible to use 100$\%$ rust
+  \item<1-> Impossible to use 100$\%$ rust - Same with C, some low level asm is needed to setup initial stack
+    pointer, etc.
   \item<2-> Rust is not as expressive as other formal verification techniques
